---
title: "First- vs Second-Order ACTT-2 SOPs with orm()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First- vs Second-Order ACTT-2 SOPs with orm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dpi = 300
)
```

```{r setup, message = FALSE, warning = FALSE}
library(markov.misc)
library(dplyr)
library(ggplot2)
library(rms)

N_PATIENTS <- 250
N_SIM <- 20
SEED <- 20260513
FOLLOW_UP <- 28
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"
MODEL_LEVELS <- c("First order: yprev", "Second order: yprev + ypprev")
```

This vignette compares two ACTT-2-like ordinal Markov models:

1. A first-order model that conditions on the previous state `yprev`.
2. A second-order model that conditions on `yprev` and the second previous
   state `ypprev`.

Both models use `rms::orm()`, patient-clustered robust covariance matrices from
`rms::robcov()`, multivariate normal coefficient simulation, and
`plot_sops()` to compare standardized state occupancy probabilities (SOPs).

`N_SIM` is intentionally small for vignette build time. Use a larger value,
such as `1000` or `2000`, for a final analysis.

## Simulate and Prepare Data

```{r simulate}
raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
model_data$y <- ordered(model_data$y, levels = STATE_LEVELS)

model_data <- model_data |>
  arrange(id, time) |>
  group_by(id) |>
  mutate(
    ypprev = dplyr::lag(yprev, default = dplyr::first(yprev))
  ) |>
  ungroup()
model_data$ypprev <- factor(model_data$ypprev, levels = levels(model_data$yprev))

baseline_data <- model_data[!duplicated(model_data$id), , drop = FALSE]

c(
  raw_rows = nrow(raw_data),
  model_rows = nrow(model_data),
  patients = length(unique(model_data$id))
)
```

`sim_actt2_brownian()` returns transition rows with the current state `y` and
previous state `yprev`. For the second-order model we add `ypprev`, the state
one step before `yprev`. The first observed transition has only one observed
history state, so this example initializes `ypprev` to the same value as
`yprev` for that transition.

```{r empirical-sops}
plot_sops(
  raw_data,
  linetype_var = "tx",
  facet_var = NULL
) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Empirical ACTT-2-like SOPs",
    x = "Day",
    y = "State occupancy probability",
    linetype = "Treatment"
  )
```

## Fit First- and Second-Order Models

The first-order model uses `yprev` only. The second-order model adds `ypprev`
to the transition model and tells `avg_sops()` to use second-order recursion via
`p2varname = "ypprev"`.

```{r fit-models}
dd <- datadist(model_data)
options(datadist = "dd")

fit_first <- orm(
  y ~ tx + rms::rcs(time, 4) + yprev,
  data = model_data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

fit_second <- orm(
  y ~ tx + rms::rcs(time, 4) + yprev + ypprev,
  data = model_data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

fit_first_robust <- robcov(fit_first, cluster = model_data$id)
fit_second_robust <- robcov(fit_second, cluster = model_data$id)
```

```{r model-size}
model_size <- data.frame(
  model = MODEL_LEVELS,
  coefficients = c(length(coef(fit_first)), length(coef(fit_second)))
)

knitr::kable(model_size)
```

## Robust MVN SOP Inference

SOPs are standardized over the same baseline patients under treatment and
control counterfactuals. The MVN simulation draws coefficients around the
`orm()` estimates using the robust covariance matrices from `robcov()`.

```{r avg-sops}
avg_first <- avg_sops(
  fit_first_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_first$yunique,
  absorb = ABSORB,
  id_var = "id"
)

avg_second <- avg_sops(
  fit_second_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  id_var = "id",
  p2varname = "ypprev"
)
```

```{r mvn-inference}
set.seed(SEED)
sops_first <- inferences(
  avg_first,
  method = "simulation",
  engine = "mvn",
  n_sim = N_SIM,
  vcov = as.matrix(fit_first_robust$var),
  conf_level = 0.95,
  return_draws = TRUE
)

set.seed(SEED)
sops_second <- inferences(
  avg_second,
  method = "simulation",
  engine = "mvn",
  n_sim = N_SIM,
  vcov = as.matrix(fit_second_robust$var),
  conf_level = 0.95,
  return_draws = TRUE
)

c(
  first_order_successful_draws = attr(sops_first, "n_successful"),
  second_order_successful_draws = attr(sops_second, "n_successful")
)
```

## Compare SOPs

```{r combine-sops}
prepare_sops_for_comparison <- function(x, model) {
  x <- as.data.frame(x)
  x$model <- model
  x$time <- as.integer(as.character(x$time))
  x
}

comparison_sops <- bind_rows(
  prepare_sops_for_comparison(sops_first, "First order: yprev"),
  prepare_sops_for_comparison(sops_second, "Second order: yprev + ypprev")
) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(as.character(state), levels = STATE_LEVELS),
    ci_width = conf.high - conf.low
  )
```

```{r sop-plot}
comparison_sops |>
  plot_sops(
    geom = "line",
    facet_var = c("tx", "model")
  ) +
  scale_x_continuous(breaks = c(1, 7, 14, 21, 28)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "First- and Second-Order SOPs with Robust MVN Intervals",
    x = "Day",
    y = "State occupancy probability",
    color = "State",
    fill = "State"
  )
```

The second-order model has more transition parameters. With a small simulated
dataset, that extra flexibility can make intervals wider even when the fitted
SOP trajectories are similar.

```{r ci-width-summary}
ci_width_summary <- comparison_sops |>
  filter(state %in% c("1", "4", "8")) |>
  group_by(model, tx, state) |>
  summarise(mean_ci_width = mean(ci_width, na.rm = TRUE), .groups = "drop")

knitr::kable(ci_width_summary, digits = 3)
```

## Compare a Derived Estimand

The stored MVN draws let us compare models on derived quantities, not just
pointwise SOPs. Here we estimate treatment minus control in expected days spent
in State 1.

```{r state1-draw-differences}
draws_first <- get_draws(sops_first)
draws_first$model <- "First order: yprev"
draws_first$time <- as.integer(as.character(draws_first$time))

draws_second <- get_draws(sops_second)
draws_second$model <- "Second order: yprev + ypprev"
draws_second$time <- as.integer(as.character(draws_second$time))

draws_all <- bind_rows(draws_first, draws_second) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    tx = as.character(tx),
    state = as.character(state)
  )

state1_time <- draws_all |>
  filter(state == "1") |>
  group_by(model, draw_id, tx) |>
  summarise(expected_days = sum(draw), .groups = "drop")

state1_control <- state1_time |>
  filter(tx == "0") |>
  select(model, draw_id, control_days = expected_days)

state1_treatment <- state1_time |>
  filter(tx == "1") |>
  select(model, draw_id, treatment_days = expected_days)

state1_diff_draws <- inner_join(
  state1_treatment,
  state1_control,
  by = c("model", "draw_id")
) |>
  mutate(delta_days = treatment_days - control_days)

state1_diff_summary <- state1_diff_draws |>
  group_by(model) |>
  summarise(
    mean_delta_days = mean(delta_days),
    lower_2.5 = unname(stats::quantile(delta_days, 0.025)),
    upper_97.5 = unname(stats::quantile(delta_days, 0.975)),
    .groups = "drop"
  )

knitr::kable(state1_diff_summary, digits = 2)
```

```{r state1-difference-plot}
ggplot(state1_diff_draws, aes(x = delta_days, y = model, fill = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey55") +
  geom_violin(trim = FALSE, alpha = 0.45, color = NA) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.85) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "MVN Draws for State 1 Treatment Difference",
    x = "Expected days in State 1, treatment minus control",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

The workflow is the same for both models through fitting and robust covariance
estimation. The second-order distinction appears in two places: the fitted
formula includes `ypprev`, and `avg_sops()` receives `p2varname = "ypprev"` so
the SOP recursion updates state-history pairs instead of only the latest state.
