---
title: "ACTT-2 SOPs with vglm and orm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACTT-2 SOPs with vglm and orm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(markov.misc)
library(dplyr)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 300
N_SIM <- 25
SEED <- 20260508
FOLLOW_UP <- 28
STATE_LEVELS <- 1:6
ABSORB <- "6"
```

This vignette compares two full proportional-odds Markov backends on the same
ACTT-2-inspired simulated data:

1. `vglm.markov()` from `markov.misc`, using `VGAM::cumulative()`.
2. `rms::orm()`, using the package's direct design-matrix fast path.

Before fitting, we collapse the ACTT-2-like states from 8 to 6 categories by
combining states 1 and 2, and states 3 and 4. The same mapping is applied to
both the current outcome `y` and previous state `yprev`. Both models then use
the same formula: treatment, a 5-knot restricted cubic spline for time, and
categorical previous state. We compute marginal state occupancy probabilities
(SOPs) with clustered score-bootstrap uncertainty, compare the model-based SOPs
with empirical SOPs, and compare the two model backends against each other.

We also fit a `vglm.markov()` partial proportional-odds extension that allows a
linear deviation from proportional odds for the linear time spline basis column.
This illustrates a VGAM-only model extension; `rms::orm()` fits full
proportional-odds models only.

`N_SIM` is intentionally small for vignette build time. Use a larger value,
for example `1000`, for a final analysis.

## Simulate And Prepare ACTT-2-Like Data

```{r simulate}
raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  seed = SEED
) |>
  mutate(
    y = case_when(
      y %in% c(1L, 2L) ~ 1L,
      y %in% c(3L, 4L) ~ 2L,
      y >= 5L ~ y - 2L
    ),
    yprev = case_when(
      yprev %in% c(1L, 2L) ~ 1L,
      yprev %in% c(3L, 4L) ~ 2L,
      yprev >= 5L ~ yprev - 2L
    )
)

data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)

baseline_data <- data[!duplicated(data$id), , drop = FALSE]

c(
  raw_rows = nrow(raw_data),
  model_rows = nrow(data),
  removed_absorbing_previous = sum(raw_data$yprev == as.integer(ABSORB)),
  remaining_absorbing_previous = sum(as.character(data$yprev) == ABSORB)
)
```

`prepare_markov_data()` removes rows where the previous state is absorbing.
The absorbing state can still appear as the current outcome, which is exactly
what we want for a transition model.

We also compute empirical arm-specific SOPs from the raw longitudinal data.
These are not model-standardized; they are a useful observed-data reference.

```{r empirical-sops}
empirical_sops <- raw_data |>
  filter(time %in% seq_len(FOLLOW_UP)) |>
  count(tx, time, state = y, name = "n") |>
  group_by(tx, time) |>
  mutate(estimate = n / sum(n)) |>
  ungroup() |>
  select(tx, time, state, estimate)

head(empirical_sops)
```

## Fit Full Proportional-Odds Markov Models

The `vglm.markov()` fit stores the metadata needed for fast SOP prediction with
inline `rms::rcs()` terms. The `orm()` fit should be created with `x = TRUE` and
`y = TRUE`, because score-bootstrap inference needs the fitted design matrix and
response.

```{r fit-models}
form_vglm <- ordered(y) ~ tx + rms::rcs(time, 5) + yprev

fit_vglm <- vglm.markov(
  form_vglm,
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = data
)

dd <- datadist(data)
options(datadist = "dd")

fit_orm <- orm(
  y ~ tx + rms::rcs(time, 5) + yprev,
  data = data,
  x = TRUE,
  y = TRUE
)

c(
  vglm_loglik = as.numeric(logLik(fit_vglm)),
  orm_loglik = as.numeric(logLik(fit_orm)),
  orm_max_deriv = unname(fit_orm$stats["Max Deriv"])
)
```

## Add A Linear-Time PPO Deviation

The full-PO model constrains every non-intercept coefficient to be the same
across all cumulative logits. A common relaxation is to keep most terms
proportional, but allow the linear component of time to have a simple
cutpoint-indexed deviation from proportional odds.

With `vglm.markov()`, we can do that by starting from the full-PO constraint
list and replacing only the linear `rms::rcs(time, 5)` basis column. The
remaining time spline basis columns, treatment, and categorical previous-state
effects stay proportional. Passing the full-PO linear predictors as `etastart`
gives the partial-PO optimizer a stable starting point.

```{r fit-linear-time-ppo}
linear_time_constraints <- fit_vglm@constraints
linear_time_term <- grep(
  "rcs\\(time, 5\\).*time$",
  names(linear_time_constraints),
  value = TRUE
)[[1]]
n_thresholds <- nrow(linear_time_constraints[[linear_time_term]])

yprev_terms <- grep("^yprev", names(linear_time_constraints), value = TRUE)
linear_time_constraints[yprev_terms] <- NULL
linear_time_constraints[["yprev"]] <- cbind(
  PO_effect = rep(1, n_thresholds)
)
linear_time_constraints[[linear_time_term]] <- cbind(
  PO_effect = rep(1, n_thresholds),
  linear_deviation = seq_len(n_thresholds) - 1
)

fit_vglm_linear_time_ppo <- vglm.markov(
  form_vglm,
  family = cumulative(reverse = TRUE, parallel = FALSE),
  data = data,
  constraints = linear_time_constraints,
  etastart = fit_vglm@predictors
)
fit_vglm_linear_time_ppo@call$constraints <- linear_time_constraints

c(
  vglm_full_po_loglik = as.numeric(logLik(fit_vglm)),
  vglm_linear_time_ppo_loglik = as.numeric(logLik(fit_vglm_linear_time_ppo))
)
```

## Score-Bootstrap SOP Uncertainty

For `vglm.markov()`, wrap the fit in `robcov_vglm()` before calling
`avg_sops()`. For `orm()`, use `rms::robcov()` for the robust covariance and
pass the row-level cluster vector to `inferences()` when using
`engine = "score_bootstrap"`.

```{r score-bootstrap}
fit_vglm_robust <- robcov_vglm(fit_vglm, cluster = data$id)
fit_vglm_linear_time_ppo_robust <- robcov_vglm(
  fit_vglm_linear_time_ppo,
  cluster = data$id
)
fit_orm_robust <- robcov(fit_orm, cluster = data$id)

avg_vglm <- avg_sops(
  fit_vglm_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = STATE_LEVELS,
  absorb = ABSORB,
  id_var = "id"
)

avg_vglm_linear_time_ppo <- avg_sops(
  fit_vglm_linear_time_ppo_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = STATE_LEVELS,
  absorb = ABSORB,
  id_var = "id"
)

avg_orm <- avg_sops(
  fit_orm_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB,
  id_var = "id"
)

set.seed(SEED)
sops_vglm <- inferences(
  avg_vglm,
  method = "simulation",
  engine = "score_bootstrap",
  n_sim = N_SIM,
  conf_level = 0.95,
  return_draws = FALSE
)

set.seed(SEED)
sops_vglm_linear_time_ppo <- inferences(
  avg_vglm_linear_time_ppo,
  method = "simulation",
  engine = "score_bootstrap",
  n_sim = N_SIM,
  conf_level = 0.95,
  return_draws = FALSE
)

set.seed(SEED)
sops_orm <- inferences(
  avg_orm,
  method = "simulation",
  engine = "score_bootstrap",
  n_sim = N_SIM,
  conf_level = 0.95,
  cluster = data$id,
  return_draws = FALSE
)
```

```{r clean-results}
clean_model_sops <- function(x) {
  as.data.frame(x) |>
    mutate(
      state = as.character(state),
      tx = as.integer(as.character(tx))
    )
}

model_sops <- bind_rows(
  list(
    "vglm.markov" = clean_model_sops(sops_vglm),
    "vglm linear-time PPO" = clean_model_sops(sops_vglm_linear_time_ppo),
    orm = clean_model_sops(sops_orm)
  ),
  .id = "backend"
) |>
  mutate(
    state = as.integer(state)
  ) |>
  select(backend, tx, time, state, estimate, conf.low, conf.high)

head(model_sops)
```

## Compare With Empirical SOPs

First, plot the empirical SOPs directly from the simulated longitudinal data.
This uses the default `plot_sops(..., geom = "line")`, so each line is the
observed proportion of patients in a state over follow-up.

```{r empirical-line-plot, fig.width = 7, fig.height = 5}
plot_sops(
  raw_data |> filter(time %in% seq_len(FOLLOW_UP)),
  facet_var = "tx"
) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    color = "State"
  )
```

The model-derived SOPs have the same line-plot structure, but add uncertainty
bands from the clustered score bootstrap. This plot includes the full-PO
`vglm.markov()` and `orm()` fits plus the VGAM linear-time PPO extension.

```{r model-line-plot, fig.width = 8, fig.height = 6}
model_sops |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = STATE_LEVELS),
    backend = factor(
      backend,
      levels = c("vglm.markov", "vglm linear-time PPO", "orm")
    )
  ) |>
  plot_sops(
    facet_var = c("tx", "backend")
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    color = "State",
    fill = "State"
  )
```

The stacked bars compare empirical arm-specific SOPs with the model-based
standardized SOPs.

```{r stacked-comparison, fig.width = 8, fig.height = 5}
stacked_sops <- bind_rows(
  empirical_sops |>
    mutate(source = "Empirical"),
  model_sops |>
    transmute(tx, time, state, estimate, source = backend)
) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = STATE_LEVELS),
    source = factor(
      source,
      levels = c(
        "Empirical",
        "vglm.markov",
        "vglm linear-time PPO",
        "orm"
      )
    )
  )

plot_sops(
  stacked_sops,
  geom = "bar",
  facet_var = c("tx", "source")
) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    title = NULL,
    x = "Day",
    y = "State occupancy probability",
    fill = "State"
  )
```

The table below summarizes absolute differences from the empirical SOPs. The
empirical rows are arm-specific while the model rows are standardized over the
baseline data, so this is a diagnostic comparison rather than a formal target.

```{r empirical-summary}
empirical_comparison <- model_sops |>
  left_join(
    empirical_sops,
    by = c("tx", "time", "state"),
    suffix = c("_model", "_empirical")
  ) |>
  group_by(backend) |>
  summarise(
    max_abs_empirical_diff = max(abs(estimate_model - estimate_empirical)),
    mean_abs_empirical_diff = mean(abs(estimate_model - estimate_empirical)),
    .groups = "drop"
  )

knitr::kable(empirical_comparison, digits = 4)
```

## Compare `vglm.markov()` And `orm()`

The two backends should produce nearly identical SOPs for this full
proportional-odds specification.

```{r backend-summary}
backend_comparison <- model_sops |>
  filter(backend %in% c("vglm.markov", "orm")) |>
  select(backend, tx, time, state, estimate) |>
  tidyr::pivot_wider(names_from = backend, values_from = estimate) |>
  mutate(diff = orm - `vglm.markov`)

backend_comparison |>
  summarise(
    max_abs_backend_diff = max(abs(diff)),
    mean_abs_backend_diff = mean(abs(diff))
  ) |>
  knitr::kable(digits = 6)
```

```{r backend-diff-plot, fig.width = 7, fig.height = 5}
backend_comparison |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = STATE_LEVELS)
  ) |>
  ggplot(aes(x = time, y = diff, color = state)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  facet_wrap(vars(tx)) +
  labs(
    x = "Day",
    y = "orm estimate minus vglm.markov estimate",
    color = "State"
  )
```

## Compare Linear-Time PPO With Full PO

The linear-time PPO model is a modest relaxation of the full-PO `vglm.markov()`
model. The difference below is the PPO SOP estimate minus the full-PO SOP
estimate for each arm, day, and state.

```{r ppo-summary}
ppo_comparison <- model_sops |>
  filter(backend %in% c("vglm.markov", "vglm linear-time PPO")) |>
  select(backend, tx, time, state, estimate) |>
  tidyr::pivot_wider(names_from = backend, values_from = estimate) |>
  mutate(diff = `vglm linear-time PPO` - `vglm.markov`)

ppo_comparison |>
  summarise(
    max_abs_ppo_diff = max(abs(diff)),
    mean_abs_ppo_diff = mean(abs(diff))
  ) |>
  knitr::kable(digits = 6)
```

```{r ppo-diff-plot, fig.width = 7, fig.height = 5}
ppo_comparison |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = STATE_LEVELS)
  ) |>
  ggplot(aes(x = time, y = diff, color = state)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  facet_wrap(vars(tx)) +
  labs(
    x = "Day",
    y = "linear-time PPO estimate minus full-PO estimate",
    color = "State"
  )
```

The first state is a useful diagnostic because it is governed by the first
cumulative boundary in these ordinal models. A linear-time PPO relaxation can
move that boundary differently over time, but it remains an ordinal model: if
state 1 is still underestimated, that is a residual model-fit issue rather than
something to solve by changing the estimand.

```{r state1-ordinal-diagnostic}
state1_ordinal_diagnostic <- model_sops |>
  filter(backend %in% c("vglm.markov", "vglm linear-time PPO", "orm")) |>
  left_join(
    empirical_sops,
    by = c("tx", "time", "state"),
    suffix = c("_model", "_empirical")
  ) |>
  filter(state == 1) |>
  mutate(diff = estimate_model - estimate_empirical) |>
  group_by(backend) |>
  summarise(
    mean_signed_state1_diff = mean(diff),
    mean_abs_state1_diff = mean(abs(diff)),
    max_abs_state1_diff = max(abs(diff)),
    .groups = "drop"
  )

knitr::kable(state1_ordinal_diagnostic, digits = 4)
```

For selected states, we can inspect the score-bootstrap intervals and overlay
the empirical SOPs as points.

```{r uncertainty-plot, fig.width = 8, fig.height = 6}
states_to_plot <- c(1, 4, 6)

plot_model_sops <- model_sops |>
  filter(state %in% states_to_plot) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = states_to_plot),
    backend = factor(
      backend,
      levels = c("vglm.markov", "vglm linear-time PPO", "orm")
    )
  )

plot_empirical_sops <- empirical_sops |>
  filter(state %in% states_to_plot) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = states_to_plot)
  )

ggplot(
  plot_model_sops,
  aes(x = time, y = estimate, color = backend, fill = backend)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.12,
    color = NA
  ) +
  geom_line(linewidth = 0.7) +
  geom_point(
    data = plot_empirical_sops,
    aes(x = time, y = estimate),
    inherit.aes = FALSE,
    shape = 21,
    fill = "white",
    color = "black",
    size = 1.6
  ) +
  facet_grid(state ~ tx) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    color = "Backend",
    fill = "Backend"
  )
```
