---
title: "ACTT-2 SOPs with Spline and Penalized Factor Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACTT-2 SOPs with Spline and Penalized Factor Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(markov.misc)
library(dplyr)
library(ggplot2)
library(rms)

N_PATIENTS <- 250
N_SIM <- 30
SEED <- 20260513
FOLLOW_UP <- 28
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"
TIME_RIDGE_PENALTY <- 100
MODEL_LEVELS <- c("RCS time, 6 knots", "Ridge-penalized factor time")
```

This vignette compares two ways to represent follow-up day in an ACTT-2-like
ordinal Markov model:

1. A numeric day term modeled as `rms::rcs(time, 6)`.
2. A saturated day factor whose indicator coefficients are shrunk with a large
   ridge penalty.

Both models use `rms::orm()` with `opt_method = "LM"`, then use
patient-clustered robust covariance matrices from `rms::robcov()` for
multivariate normal simulation intervals around standardized SOPs.

`N_SIM` is intentionally small for vignette build time. Use a larger value,
for example `1000` or `2000`, for a final analysis.

## Simulate and Prepare Data

```{r simulate}
raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  seed = SEED
)

model_data_rcs <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
model_data_rcs$y <- ordered(model_data_rcs$y, levels = STATE_LEVELS)

model_data_factor <- model_data_rcs
model_data_factor$time <- factor(
  model_data_factor$time,
  levels = as.character(seq_len(FOLLOW_UP))
)

baseline_rcs <- model_data_rcs[!duplicated(model_data_rcs$id), , drop = FALSE]
baseline_factor <- model_data_factor[
  !duplicated(model_data_factor$id),
  ,
  drop = FALSE
]

c(
  raw_rows = nrow(raw_data),
  model_rows = nrow(model_data_rcs),
  patients = length(unique(model_data_rcs$id))
)
```

The spline model keeps `time` numeric. The factor model uses the same rows and
same previous-state coding, but turns `time` into the ordered set of scheduled
daily visits.

## Fit the Time Models

The factor model has 27 visit indicators before penalization. We first fit the
unpenalized factor model only to get its expanded design matrix, then create a
ridge penalty matrix that is zero for treatment and previous-state effects and
positive only for the factor-time columns.

```{r fit-models}
dd_rcs <- datadist(model_data_rcs)
options(datadist = "dd_rcs")

fit_rcs <- orm(
  y ~ tx + rms::rcs(time, 6) + yprev,
  data = model_data_rcs,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

dd_factor <- datadist(model_data_factor)
options(datadist = "dd_factor")

fit_factor_template <- orm(
  y ~ tx + time + yprev,
  data = model_data_factor,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

time_cols <- grep("^time=", colnames(fit_factor_template$x))
time_penalty <- matrix(
  0,
  nrow = ncol(fit_factor_template$x),
  ncol = ncol(fit_factor_template$x),
  dimnames = list(
    colnames(fit_factor_template$x),
    colnames(fit_factor_template$x)
  )
)
time_penalty[time_cols, time_cols] <- diag(length(time_cols))

fit_factor <- orm(
  y ~ tx + time + yprev,
  data = model_data_factor,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100,
  penalty = TIME_RIDGE_PENALTY,
  penalty.matrix = time_penalty
)
```

The factor-time model still lets every follow-up day have its own coefficient,
but the large ridge penalty reduces the effective time degrees of freedom to a
size similar to the 6-knot spline basis.


```{r effective-df}
fit_rcs
fit_factor
```


## Robust MVN SOP Uncertainty

For both models, SOP point estimates are standardized over the same baseline
patients under treatment and control counterfactuals. The MVN simulation uses
the patient-clustered robust covariance matrix from `rms::robcov()`.

```{r robust-vcov}
fit_rcs_robust <- robcov(fit_rcs, cluster = model_data_rcs$id)
fit_factor_robust <- robcov(fit_factor, cluster = model_data_factor$id)

vcov_rcs_robust <- as.matrix(fit_rcs_robust$var)
vcov_factor_robust <- as.matrix(fit_factor_robust$var)
```

```{r avg-sops}
avg_rcs <- avg_sops(
  fit_rcs_robust,
  newdata = baseline_rcs,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_rcs$yunique,
  absorb = ABSORB,
  id_var = "id"
)

avg_factor <- avg_sops(
  fit_factor_robust,
  newdata = baseline_factor,
  variables = list(tx = c(0, 1)),
  times = NULL,
  ylevels = fit_factor$yunique,
  absorb = ABSORB,
  id_var = "id"
)
```

```{r mvn-inference}
set.seed(SEED)
sops_rcs <- inferences(
  avg_rcs,
  method = "simulation",
  engine = "mvn",
  n_sim = N_SIM,
  vcov = vcov_rcs_robust,
  conf_level = 0.95,
  return_draws = TRUE
)

set.seed(SEED)
sops_factor <- inferences(
  avg_factor,
  method = "simulation",
  engine = "mvn",
  n_sim = N_SIM,
  vcov = vcov_factor_robust,
  conf_level = 0.95,
  return_draws = TRUE
)

c(
  rcs_successful_draws = attr(sops_rcs, "n_successful"),
  factor_successful_draws = attr(sops_factor, "n_successful")
)
```

## Compare SOPs

```{r combine-sops}
prepare_sops_for_comparison <- function(x, model) {
  x <- as.data.frame(x)
  x$model <- model
  x$time <- as.integer(as.character(x$time))
  x
}

comparison_sops <- bind_rows(
  prepare_sops_for_comparison(sops_rcs, "RCS time, 6 knots"),
  prepare_sops_for_comparison(
    sops_factor,
    "Ridge-penalized factor time"
  )
) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(as.character(state), levels = STATE_LEVELS),
    ci_width = conf.high - conf.low
  )
```

```{r selected-sops-plot}
comparison_sops |>
  # filter(state %in% c("1", "4", "8")) |>
  plot_sops(
    geom = "line",
    facet_var = c("tx", "model"),
    ribbon_alpha = 0.15
  ) +
  scale_x_continuous(breaks = c(1, 7, 14, 21, 28)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Selected SOPs with Robust MVN Intervals",
    x = "Day",
    y = "State occupancy probability",
    color = "State",
    fill = "State"
  )
```

Results are very similar, but the factor-time model seems to haver larger uncertainty.

## Contrasts on derived estimands

We can also use the stored MVN draws to compare a derived estimand: treatment
minus control in expected days spent in State 1. Here expected days are computed
by summing the daily State 1 SOP draws over follow-up.

```{r state1-draw-differences}
draws_rcs <- get_draws(sops_rcs)
draws_rcs$model <- "RCS time, 6 knots"
draws_rcs$time <- as.integer(as.character(draws_rcs$time))

draws_factor <- get_draws(sops_factor)
draws_factor$model <- "Ridge-penalized factor time"
draws_factor$time <- as.integer(as.character(draws_factor$time))

draws_all <- bind_rows(draws_rcs, draws_factor) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    tx = as.character(tx),
    state = as.character(state)
  ) |> 
  arrange(draw_id, time, tx)

state1_time <- draws_all |>
  filter(state == "1") |>
  group_by(model, draw_id, tx) |>
  summarise(expected_days = sum(draw), .groups = "drop")

state1_control <- state1_time |>
  filter(tx == "0") |>
  select(model, draw_id, control_days = expected_days)

state1_treatment <- state1_time |>
  filter(tx == "1") |>
  select(model, draw_id, treatment_days = expected_days)

state1_diff_draws <- inner_join(
  state1_treatment,
  state1_control,
  by = c("model", "draw_id")
) |>
  mutate(delta_days = treatment_days - control_days)

state1_diff_summary <- state1_diff_draws |>
  group_by(model) |>
  summarise(
    mean_delta_days = mean(delta_days),
    lower_2.5 = unname(stats::quantile(delta_days, 0.025)),
    upper_97.5 = unname(stats::quantile(delta_days, 0.975)),
    .groups = "drop"
  )

knitr::kable(state1_diff_summary, digits = 2)
```

```{r state1-difference-plot}
ggplot(state1_diff_draws, aes(x = delta_days, y = model, fill = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey55") +
  geom_violin(trim = FALSE, alpha = 0.45, color = NA) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.85) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "MVN Draws for State 1 Treatment Difference",
    x = "Expected days in State 1, treatment minus control",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```