---
title: "Analytical Confidence Intervals for State Occupancy Probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analytical Confidence Intervals for State Occupancy Probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

`inferences(method = "delta")` adds analytical confidence intervals to
estimated state occupancy probabilities (SOPs): the probabilities of being in
each state at a given time. It calculates standard errors from how these
probabilities change with the fitted model coefficients. With unconditional
variance, it also accounts for variation in which patients were sampled.

You must select `method = "delta"` explicitly. The default is `method = "mvn"`,
which simulates model coefficients to calculate confidence intervals.

## Fit a Model

This example simulates patient data and fits a proportional-odds model using
`orm_markov()`. Setting `id_var = "id"` identifies observations from the same
patient, so the variance calculation can account for their dependence.

```{r setup}
library(markov.misc)
library(rms)
library(patchwork)
library(ggplot2)

N_PATIENTS <- 800
FOLLOW_UP <- 30
SEED <- 20260718
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_prob = 0.5,
  absorbing_state = as.integer(ABSORB),
  seed = SEED,
  mu_treatment_effect = -0.5
)

plot_sops(raw_data, linetype_var = "tx", facet_var = NULL)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB)
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit <- orm_markov(
  y ~ tx + time + yprev,
  data = model_data,
  id_var = "id",
  type = "HC0",
  cadjust = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)
```

## Choose Conditional or Unconditional Variance

Use `inferences(method = "delta", vcov = "conditional")` to account for
uncertainty in the estimated model coefficients while treating the patients'
starting states and covariates used for prediction as fixed. For an average,
this answers how precisely we have estimated the average for those patients.

Use `vcov = "unconditional"` to also account for variation in which patients
are sampled from the population. This estimates uncertainty in the population
average, including the association between the sampled patients and the model
coefficients estimated from them. Both choices give the same point estimates;
their standard errors and confidence intervals can differ.

Averaged SOPs and comparisons default to unconditional variance; individual
`sops()` results support only conditional variance.

For conditional variance, let $h_i(\theta)$ contain patient $i$'s predicted
probabilities and $G_i$ their derivatives with respect to the model coefficients
$\theta$. The derivative of the average is $\bar G = n^{-1}\sum_i G_i$.
With coefficient covariance $V_\theta$, the covariance of the average is
$\bar G V_\theta \bar G^{\mathsf T}$.

By default, $V_\theta$ is the patient-cluster robust covariance from the fit.
For `orm_markov()` and `vglm_markov()`, `type = "HC1"` multiplies that
covariance by $(m - 1) / (m - p)$, where $m$ is the number of fitting rows
(positive-weight rows for ORM) and $p$ is the number of coefficients.
`cadjust = TRUE` independently multiplies it by $n / (n - 1)$, where $n$ is
the number of patient clusters. Clustered wrapper fits enable `cadjust` by
default. Conditional analytical variance and MVN inference also support
weighted and penalized ORM fits when their fitted covariance is valid.

```{r conditional-variance}
t1 <- Sys.time()
avg_conditional <- avg_sops(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |> inferences(
  method = "delta",
  vcov = "conditional",
  conf_type = "auto"
)
t2 <- Sys.time()
t2 - t1

comp_conditional <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB,
  estimand = "time_in_state",
  comparison = "difference"
) |>
  inferences(
    method = "delta",
    vcov = "conditional"
  )

p1.sops <- plot_sops(avg_conditional, linetype_var = "tx", facet_var = NULL) + labs(
  title = "Conditional Variance"
)

p1.c <- plot_comparisons(comp_conditional) + labs(
  title = "Conditional Variance"
)
```

For unconditional variance, the calculation combines variation between
patients' predictions with their contributions to coefficient estimation.
Each patient contributes the vector

$$
\phi_i = h_i - \bar h + \bar G A^{-1}s_i,
$$

where $n$ is the number of patients, $\bar h$ is their average prediction,
$s_i$ is the sum of the derivatives of patient $i$'s log likelihood with
respect to the coefficients, and $A$ is the negative derivative of the average
patient score. Thus $A^{-1}s_i$ describes patient $i$'s contribution to
coefficient estimation. It uses `stats::cov(phi) / n`, retaining the
association between these two contributions in the observed sample.
Here `stats::cov()` uses the usual sample-covariance divisor, $n - 1$.
The fitting options `type` and `cadjust` do not apply an additional correction
to this unconditional variance estimate.

```{r unconditional-variance}
t1 <- Sys.time()
avg_unconditional <- avg_sops(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |> inferences(
  method = "delta",
  vcov = "unconditional",
  conf_type = "wald"
)
t2 <- Sys.time()
t2 - t1

comp_unconditional <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB,
  estimand = "time_in_state",
  comparison = "difference"
) |>
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

p2.sops <- plot_sops(avg_unconditional, linetype_var = "tx", facet_var = NULL) + labs(title = "Unconditional Variance")
p2.c <- plot_comparisons(comp_unconditional) + labs(
  title = "Unconditional Variance"
)
```

```{r}
p1.sops + p2.sops + plot_layout(guides = 'collect')
p1.c + p2.c
```

Unconditional variance requires the patients stored with the fitted model,
with one complete starting state and set of covariates per patient and matching
patient IDs in the fitting data. It is unavailable for user-supplied `newdata`.
The calculation uses unadjusted likelihood scores and model-based sensitivity,
with the patient sample-covariance correction described above. The fitting
options `type` and `cadjust` do not change it. Weighted and penalized ORM fits
support conditional variance only.

Use a model fitted with `orm_markov()` or `vglm_markov()` for analytical
confidence intervals. These functions store the patient information needed
for the calculations. Models fitted directly with `rms::orm()` or
`VGAM::vglm()` cannot be used in this workflow.

### Which Patients Are Included in the Average?

When you omit `newdata`, predictions use each fitted patient's starting state
and covariates at `first_followup_time`. The model stores this information
before dropping rows with missing outcomes.

- Each patient must have exactly one row at that time with their ID,
  predictors, time, and previous state (`yprev`) observed.
- The outcome on that row may be missing if the patient has another usable
  transition in the data used to fit the model.

For numeric time, `first_followup_time` defaults to 1. Time 1 must be present,
and the schedule cannot contain earlier values. If you specify another value,
it must be the earliest scheduled follow-up. For factor or character time,
you must specify a matching value explicitly. A later observation cannot
replace missing starting information.

These variance estimates do not correct bias caused by missing observations
that depend on unobserved patient outcomes.

Individual `sops()` results default to `vcov = "conditional"` and cannot use
`"unconditional"`. Averaged SOPs and comparisons default to `"unconditional"`.
A supplied coefficient covariance matrix selects conditional inference for any
of these result types. Omitted `vcov` and `NULL` are equivalent. These string
choices are available only with `method = "delta"`.
Unsupported unconditional inference errors without falling back, so supplied
`newdata` requires an explicit conditional choice.

## Why Patient Clustering Matters

A patient can contribute several observations to the model. Treating those
observations as independent can give misleading standard errors. The package
uses patient IDs to account for dependence within each patient:

1. If you supply `cluster`, its entries must identify the patient for each row
   used to fit the model, in the same order.
2. Otherwise, the package uses the ID column named by `id_var` when fitting
   with `orm_markov()` or `vglm_markov()`.
3. Without patient IDs, it cannot calculate patient-cluster robust variance.

For conditional variance, you can instead supply a coefficient covariance
matrix through `vcov`. It must include all model coefficients, with matching
names. This selects conditional variance: a covariance matrix alone does not
provide the individual patient information required for unconditional variance.

Accounting for dependence within patients does not correct an inappropriate
transition model or bias from missing observations.

## Logit and Wald Intervals

For probabilities, `conf_type = "auto"` calculates the interval on the logit
scale and transforms its limits back to probabilities. This keeps the limits
between zero and one. Each interval applies to one estimate; the stated
confidence level does not apply to all intervals jointly.

If an estimated probability is exactly zero or one, the logit is undefined.
The function returns `NA` limits and a warning, even if the standard error is
zero. Such values can arise from numerical rounding and do not necessarily
mean that the outcome is certain. An absorbing state does not change this:
although patients stay there once they enter it, the probability of reaching
it still has to be estimated.

`conf_type = "wald"`, used in the examples above, calculates the estimate plus
or minus a normal quantile times its standard error. Its limits can fall
outside zero and one. Analytical intervals for differences always use this
Wald calculation; `conf_type = "auto"` selects it automatically.

```{r comparison}
sop_difference <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  estimand = "sop",
  state_sets = c("1", "2"),
  comparison = "difference",
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "conditional",
    conf_type = "auto"
  )

head(sop_difference)
```

The example subtracts the average probability under the first treatment value
from that under the second. The standard error accounts for the fact that
both predictions come from the same fitted model and patients.

You can also compare expected time in a state with `estimand = "time_in_state"`.
For visits recorded as factors, supply the elapsed time corresponding to each
visit. See the factor-time vignette for examples of specifying those times
and including the observed baseline state.

## Extract Variances and Covariances

The output from `inferences()` already includes standard errors and confidence
limits. Use `vcov()` if you also need to know how uncertainty in different
estimates is related. Its diagonal entries are the variances of the estimates;
the other entries are covariances between pairs of estimates.

The example below requests the first six estimates. The returned covariance
uses whichever choice, conditional or unconditional, you made when calling
`inferences()`. Taking the square roots of its diagonal gives the standard
errors for those rows.

```{r accessors}
rows <- seq_len(min(6L, nrow(avg_conditional)))

conditional_covariance <- stats::vcov(avg_conditional, rows = rows)
unconditional_covariance <- stats::vcov(avg_unconditional, rows = rows)

sqrt(diag(conditional_covariance))
sqrt(diag(unconditional_covariance))
```

For readers interested in the derivatives used in the calculation, the
internal function below shows how each selected estimate changes with each
model coefficient. Each row corresponds to an estimate and each column to a
coefficient. These derivatives alone are not standard errors and do not
include the additional patient-sampling contribution to unconditional variance.
Use `inferences()` and `vcov()` for the uncertainty calculations.

```{r coefficient-derivatives}
coefficient_derivatives <- markov.misc:::get_jacobian(avg_conditional, rows = rows)
coefficient_derivatives
```

Without `rows`, `vcov()` returns a covariance for every pair of estimates.
This can require substantial memory: doubling the number of estimates
quadruples the number of entries. Request only the rows you need for large
results. The option `markov.misc.delta_max_bytes`, which defaults to 256 MiB,
limits the main memory allocations used by analytical inference and covariance
extraction. It is not a limit on the total memory used by R.

## Supported Models and Comparisons

Analytical confidence intervals are available for:

- first-order Markov models, where the next state depends on the previous
  state and the model covariates;
- full proportional odds with a reverse cumulative-logit link;
- frequentist models created by `orm_markov()` or `vglm_markov()` (including
  the `robcov_vglm` object returned by the latter);
- individual SOPs with conditional variance, and average SOPs with conditional
  or unconditional variance;
- difference comparisons for `estimand = "sop"` and
  `estimand = "time_in_state"`; and
- numeric and factor visit times.

The following requests are not supported and return an error:

| Request | Current status |
| --- | --- |
| Individual `sops()` with `vcov = "unconditional"` | Unsupported; omit `vcov` or use `vcov = "conditional"`. |
| Ratio comparison | Use differences for analytical intervals. |
| `estimand = "time_benefit"` | Analytical intervals are not available for this comparison. |
| Analytical intervals for groups specified with `by` | Not currently supported. |
| Partial/nonproportional odds | Only full proportional-odds models are supported. |
| Analytical intervals for second-order Markov models | Use simulation or bootstrap inference where supported. |
| Analytical intervals for `blrm` models | Unsupported; use the model's posterior SOP intervals. |
| Unconditional variance with user `newdata` | Unsupported; use `vcov = "conditional"`. |
| Unconditional variance with a weighted or penalized ORM fit | Use conditional variance. |
| Random or independent external population | Intentionally unsupported and out of scope. |
| Intervals intended to cover all estimates jointly | The confidence level applies to each interval separately. |

Analytical intervals use an approximation based on small changes in the model
coefficients. Simulation and bootstrap intervals can differ because they
recalculate predictions over larger changes in coefficients or refitted data.
