---
title: "Analytical Confidence Intervals for State Occupancy Probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analytical Confidence Intervals for State Occupancy Probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

`inferences(method = "delta")` adds deterministic first-order confidence
intervals to state occupancy probabilities (SOPs) from supported frequentist
Markov models. Unlike multivariate-normal simulation or bootstrap inference,
the delta method does not generate draws. It differentiates the first-order SOP
recursion analytically, then propagates coefficient or patient-level influence
uncertainty.

The existing default remains `method = "mvn"`. Choose `method = "delta"`
explicitly when the analytical target and supported model scope described below
are appropriate.

## A Small Full Proportional-Odds Example

The package-aware fitting wrapper stores row-aligned fitting data and patient ID
metadata. Those IDs define the independent clusters for analytical covariance
and superpopulation influence calculations.

```{r setup}
library(markov.misc)
library(rms)

N_PATIENTS <- 80
FOLLOW_UP <- 6
SEED <- 20260718
ABSORB <- "6"

raw_data <- sim_trajectories_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_prob = 0.5,
  absorbing_state = as.integer(ABSORB),
  seed = SEED,
  mu_treatment_effect = -0.03
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB)
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit <- orm_markov(
  y ~ tx + time + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

avg <- avg_sops(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
)
```

## Choose the Estimand

For an empirical target, the observed standardization profiles are held fixed.
If $h_i(\theta)$ is patient $i$'s vector of counterfactual SOP cells and
$G_i(\theta)$ is its coefficient Jacobian, the average Jacobian is

$$
\bar G = n^{-1}\sum_i G_i.
$$

With coefficient covariance $V_\theta$, the empirical-target covariance is
$\bar G V_\theta \bar G^{\mathsf T}$. By default, $V_\theta$ is the
patient-cluster robust covariance associated with the fit.

```{r empirical-target}
avg_empirical <- inferences(
  avg,
  method = "delta",
  target = "empirical",
  conf_type = "auto"
)

head(avg_empirical)
```

For a superpopulation target, the fitted cohort represents a sample from a
superpopulation. Its profile distribution therefore also contributes
uncertainty. The implementation constructs one influence row per patient,

$$
\phi_i = h_i - \bar h + \bar G A^{-1}s_i,
$$

where $s_i$ is the patient-aggregated likelihood score and $A^{-1}$ is inverse
per-patient sensitivity. It uses `stats::cov(phi) / n`, retaining the
finite-sample cross-covariance between profile and model-score contributions.
Because `stats::cov()` divides by $n - 1$, this equals the centered
patient-level HC0 crossproduct multiplied by $n / (n - 1)$. That correction is
part of the superpopulation estimator; it is not the fitted backend's row-level
HC1 or cluster adjustment.

```{r superpopulation-target}
avg_superpopulation <- inferences(
  avg,
  method = "delta",
  target = "superpopulation",
  conf_type = "auto"
)

head(avg_superpopulation)
```

The superpopulation route is a fitted-cohort calculation. It requires one
stored first-follow-up profile per fitted patient and exact patient-ID
alignment with the fitted likelihood scores. It is not an external-target or
random-effects population analysis. VGLM `type` and `cadjust` choices affect the
empirical coefficient sandwich, but the stacked superpopulation calculation uses
its own patient sample-covariance convention and reports the ignored backend
settings in covariance metadata.

### Which patients define the fitted cohort?

The fitting wrappers retain starting-profile information before a missing
transition response can remove a row from the model frame. The contract is:

- a fitted patient must have exactly one complete profile at the designated
  `first_followup_time`, containing ID, treatment/covariates, modeled time
  information, and the known previous state such as `yprev`;
- the transition response at that row is not required;
- the patient must contribute at least one usable likelihood transition
  somewhere in the fitted data;
- a complete starting profile plus a missing first outcome plus a later usable
  transition is included;
- a starting profile with no usable transition anywhere is excluded; and
- likelihood rows without a complete first-follow-up profile are an error.

`first_followup_time` only selects and validates this automatic profile. For
numeric time, its `NULL` default resolves to 1: time 1 must be present and the
schedule may not contain values below 1. An explicit numeric value must be the
earliest scheduled follow-up. Factor or character time requires an explicit
matching value. The wrappers no longer accept `start_time`, `origin_time`, or
`time_map`; real-time mapping and baseline anchoring belong to downstream SOP
summaries. Later likelihood rows never silently replace a missing designated
profile. This standardization contract assumes the observation process is
non-informative; the analytical variance does not correct informative
missingness.

Individual `sops()` results accept only an omitted target or
`target = "fixed"`. Empirical and superpopulation targets apply only to averaged
SOPs and average comparisons, which use `target = "empirical"` by default. The
unreleased `target = "population"` spelling is not accepted. Supplying the target
explicitly is helpful whenever the distinction matters scientifically.

## Why Patient Clustering Matters

Transition models contain repeated rows per patient. Those rows are not
independent clusters. Analytical inference therefore follows this invariant:

1. An explicit row-aligned `cluster` vector takes precedence.
2. Otherwise, the model must store both its fitting data and `id_var` metadata,
   as `orm_markov(id_var = ...)` and `vglm_markov(id_var = ...)` do.
3. Observation rows are never silently substituted as clusters.

For an empirical target, a complete user-supplied coefficient covariance can
replace the fitted covariance after strict dimension, name, finiteness,
symmetry, and positive-semidefiniteness checks. A custom covariance is not
accepted for `target = "superpopulation"`, because a covariance matrix alone does
not contain the joint score/profile information needed by the stacked influence
function.

Patient-cluster robustness protects the variance calculation against arbitrary
within-patient score correlation. It does not remove bias from a misspecified
transition model, an invalid Markov assumption, informative observation, or an
incorrect proportional-odds restriction.

## Logit and Wald Intervals

For SOP probabilities, `conf_type = "auto"` selects componentwise logit-delta
intervals. They remain inside $[0, 1]$, but they are not a simultaneous region
for the probability simplex. Exact structural boundary cells with zero
standard error receive degenerate intervals; a nonstructural boundary where the
logit derivative is undefined is reported as `NA` with a warning.

`conf_type = "wald"` instead uses identity-scale normal intervals and may extend
outside $[0, 1]$. Supported average comparisons always use Wald intervals;
`conf_type = "auto"` selects them automatically.

```{r comparison}
sop_difference <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  estimand = "sop",
  state_sets = c("1", "2"),
  comparison = "difference",
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    target = "empirical",
    conf_type = "auto"
  )

head(sop_difference)
```

Analytical differences are linear transformations of average SOP cells.
`estimand = "time_in_state"` is also supported, including factor-visit designs
with real-time mapping, interpolation, an observed baseline anchor, and
trapezoidal integration. See the factor-time vignette for that workflow.

## Inspect the Analytical State

Delta results do not store simulation draws or a dense covariance across every
result cell. They retain a factorized analytical representation:

- fixed individual and empirical averaged results store a result-cell Jacobian
  and coefficient covariance;
- superpopulation results store the average Jacobian and patient influence
  matrix.

Use `rows` to materialize only the block needed downstream.

```{r accessors}
rows <- seq_len(min(6L, nrow(avg_empirical)))

jacobian_block <- get_jacobian(avg_empirical, rows = rows)
empirical_covariance <- stats::vcov(avg_empirical, rows = rows)
superpopulation_covariance <- stats::vcov(avg_superpopulation, rows = rows)

dim(jacobian_block)
diag(empirical_covariance)
diag(superpopulation_covariance)
```

Numeric, logical, and stored character row keys are accepted. Analytical
workspace, retained state, and accessor materialization are guarded by the
`markov.misc.delta_max_bytes` option, which defaults to 256 MiB.

## Supported Scope and Explicit Errors

The current analytical implementation supports:

- first-order Markov recursion;
- full proportional odds with a reverse cumulative-logit link;
- frequentist `orm`, `vglm`, and `robcov_vglm` models;
- individual SOPs, empirical average SOPs, and fitted-cohort superpopulation
  average SOPs;
- difference comparisons for `estimand = "sop"` and
  `estimand = "time_in_state"`; and
- numeric and factor visit designs that can be represented by the first-order
  execution plan.

The following requests fail explicitly rather than switching methods or
estimands:

| Request | Current status |
| --- | --- |
| Individual `sops()` with an empirical or superpopulation target | Unsupported; omit `target` or use `target = "fixed"`. |
| Ratio comparison | Planned follow-up; analytical comparisons currently support differences only. |
| `estimand = "time_benefit"` | Planned follow-up because it is nonlinear in paired state distributions. |
| `by`-stratum analytical inference | Planned follow-up. |
| Partial/nonproportional odds | Planned follow-up outside the current full-PO derivative scope. |
| Second-order delta recursion | Planned follow-up; use an existing draw or refit method when applicable. |
| `blrm` delta inference | Unsupported; use the model's posterior SOP intervals. |
| Superpopulation target with user `newdata` | Unsupported; supplied profiles are fixed. |
| Superpopulation target with a weighted or penalized ORM fit | Unsupported until the corresponding score/sensitivity contract is established; empirical delta inference remains available. |
| Random or independent external population | Intentionally unsupported and out of scope. |
| Simultaneous confidence bands or simplex regions | Out of scope; intervals are componentwise. |

Analytical gradients are checked against central finite differences in the test
suite. Production inference always uses the analytic recursion. MVN and
bootstrap procedures remain useful alternatives, but numerical agreement with
their nonlinear draw distributions is not a validation requirement for this
first-order approximation.
