---
title: "Bayesian SOPs with blrm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian SOPs with blrm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
RUN_BAYESIAN <- identical(
  Sys.getenv("MARKOV_MISC_RUN_BAYESIAN_VIGNETTE"),
  "true"
) && requireNamespace("rmsb", quietly = TRUE)
```

`Markov.misc` also supports Bayesian transition models fitted with `blrm_markov()`, a stored-data wrapper around `rmsb::blrm()`. In this vignette we use a small number of MCMC iterations and SOP draws for demonstration.

## Data

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)
library(rmsb)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260601
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"
SOP_DRAWS <- 200

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(raw_data, absorbing_state = 8)
```

As usual, we plot the raw data.

```{r}
plot_sops(raw_data)
```

## Fit `blrm_markov()`

The `ppo = ~ time` term marks `time` as non-proportional. The `cppo` function
constrains that deviation to be linear in the numeric ordinal score.

```{r fit-blrm, eval = RUN_BAYESIAN}
dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_blrm <- blrm_markov(
  y ~ yprev + rms::rcs(time, 4) * tx + cluster(id),
  ppo = ~ time,
  cppo = function(y) as.numeric(y),
  data = model_data,
  id_var = "id",
  iter = 300,
  warmup = 150,
  chains = 2,
  method = "sampling",
  refresh = 0,
  seed = SEED
)
```

## Posterior SOPs

For Bayesian fits, `avg_sops()` evaluates the SOP recursion on posterior draws.
There is no separate `inferences()` step. [add section on options / chunking the draws]

```{r avg-sops, eval = RUN_BAYESIAN}
options(markov.misc.blrm_avg_chunk_size = 100)

sop_blrm <- avg_sops(
  fit_blrm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_blrm$ylevels,
  absorb = ABSORB,
  include_re = TRUE,
  n_draws = NULL,
  seed = SEED,
  posterior_summary = "mean",
  conf_level = 0.90,
  return_draws = TRUE
)

head(sop_blrm)
```

```{r plot-sops, eval = RUN_BAYESIAN}
plot_sops(
  sop_blrm,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "Posterior state occupancy probability"
  )
```

## Model Diagnostics

These diagnostics compare empirical trajectory features with features implied
by the fitted transition model. For `blrm_markov()` plots, posterior draws are
kept separate through the transition, correlation, and variogram summaries;
`seed` controls posterior-draw selection.

```{r diagnostics-transitions-observed, eval = RUN_BAYESIAN}
diagnostic_times <- c(1, 7, FOLLOW_UP)

plot_transitions(
  raw_data,
  times = diagnostic_times,
  ylevels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model, eval = RUN_BAYESIAN}
plot_transitions(
  fit_blrm,
  times = diagnostic_times,
  ylevels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED,
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed, eval = RUN_BAYESIAN}
plot_correlation(
  raw_data,
  times = seq_len(FOLLOW_UP),
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model, eval = RUN_BAYESIAN}
plot_correlation(
  fit_blrm,
  times = seq_len(FOLLOW_UP),
  ylevels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED,
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed, eval = RUN_BAYESIAN}
plot_variogram(
  raw_data,
  times = seq_len(FOLLOW_UP)
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model, eval = RUN_BAYESIAN}
plot_variogram(
  fit_blrm,
  times = seq_len(FOLLOW_UP),
  ylevels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED
) +
  labs(title = "Model-based correlation variogram")
```
