---
title: "Marginal SOPs with rmsb::blrm()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal SOPs with rmsb::blrm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

This vignette shows the basic workflow for using a Bayesian ordinal Markov
model fit with `rmsb::blrm()` in `avg_sops()`. The example uses a first-order
Markov model and includes a patient random effect via `cluster(id)`.

The model-fitting chunks are not evaluated when the vignette is built, because
Stan sampling requires a configured backend and can take several minutes. The
numbers below are intentionally small for an example: 500 patients, short Stan
chains, and only 25 posterior draws for the SOP calculation. Increase these for
an analysis.

## Setup

```{r setup, eval = FALSE}
library(markov.misc)
library(ggplot2)
library(rms)
library(rmsb)

N_PATIENTS <- 200
FOLLOW_UP <- 28
STATE_LEVELS <- 1:8
ABSORB <- "8"
SEED <- 20260512
SOP_DRAWS <- 25
```

## Simulate ACTT-2-Like Brownian Data

`sim_actt2_brownian()` creates ACTT-2-inspired ordinal trajectories. We keep one
row per patient for standardization and use the transition-level data for model
fitting.

```{r data, eval = FALSE}
brownian_actt2 <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_prob = 0.5,
  seed = SEED
)

model_data <- prepare_markov_data(
  brownian_actt2,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)

baseline_data <- model_data[!duplicated(model_data$id), , drop = FALSE]

c(
  patients = length(unique(brownian_actt2$id)),
  model_rows = nrow(model_data),
  baseline_rows = nrow(baseline_data)
)
```

## Fit A First-Order blrm Model

The previous state enters through `yprev`, so this is a first-order Markov
model. The `cluster(id)` term creates fitted patient random effects. Those
effects can be used later in prediction with `include_re = TRUE` as long as the
prediction data contain known fitted IDs.

```{r fit, eval = FALSE}
dd <- datadist(model_data)
options(datadist = "dd")
options(mc.cores = 2)

fit_blrm <- blrm(
  y ~ tx + rcs(time, 4) + yprev + cluster(id),
  data = model_data,
  method = "sampling",
  backend = "cmdstan",
  chains = 2,
  iter = 500,
  refresh = 0,
  loo = FALSE,
  seed = SEED
)
```

## Compute Marginal SOPs

For `blrm()` fits, `avg_sops()` samples posterior draws and computes the SOPs on
those draws directly. The `estimate` column is the requested posterior summary
of the SOP draws, not a plug-in calculation from summarized model parameters.
If you use `posterior_summary = "median"`, the medians are computed
state-by-state, so their displayed totals are not guaranteed to sum to exactly
one. The draw-level probabilities remain normalized; use
`posterior_summary = "mean"` when the displayed estimates must preserve total
probability.

Here `include_re = TRUE` adds the fitted patient random effects to the
predictions. Use `include_re = FALSE` for new IDs or for predictions that omit
patient effects.

```{r avg-sops, eval = FALSE}
avg_blrm <- avg_sops(
  fit_blrm,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_blrm$ylevels,
  absorb = ABSORB,
  id_var = "id",
  include_re = TRUE,
  n_draws = SOP_DRAWS,
  seed = SEED,
  posterior_summary = "mean",
  conf_level = 0.90,
  return_draws = TRUE
)

head(avg_blrm)
```

There is no separate `inferences()` step for `blrm` SOPs. Posterior uncertainty
has already been computed by `avg_sops()`. To change the interval level or the
posterior draws, rerun `avg_sops()` with different `conf_level`, `n_draws`, or
`seed`.

## Plot The SOPs

`plot_sops()` reads the posterior interval columns and draws uncertainty
ribbons for line plots.

```{r line-plot, eval = FALSE}
plot_sops(
  avg_blrm,
  geom = "line",
  facet_var = "tx"
) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Posterior Marginal State Occupancy Probabilities",
    x = "Day",
    y = "State occupancy probability"
  )
```

If `return_draws = TRUE`, stacked bar plots can overlay a small subset of
posterior draw-specific stacks.

```{r bar-plot, eval = FALSE}
plot_sops(
  avg_blrm,
  geom = "bar",
  facet_var = "tx",
  n_draws = 8,
  draw_alpha = 0.08
) +
  labs(
    title = "Posterior Draws Behind The Marginal SOPs",
    x = "Day",
    y = "State occupancy probability"
  )
```
