---
title: "Bootstrap Marginal State Occupancy Probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap Marginal State Occupancy Probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  fig.align = "center",
  dpi = 300,
  fig.retina = 2        
)
```

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(markov.misc)
library(dplyr)
library(tidyr)
library(rms)
library(VGAM)
library(ggplot2)
library(ggdist)
library(patchwork)

N_PATIENTS <- 800
R <- 50 # Reduced for vignette build time; use higher (e.g., 1000) for analysis
SEED <- 2375679
FU <- 30
```

## Step 1: Simulate Trial Data

We simulate a discrete-time process using a latent Brownian motion model.
We also add a baseline age covariate (one value per patient) drawn from a
normal distribution. Age is not part of the outcome-generating mechanism, but
we include it in the fitted models to illustrate standardization over an
additional baseline covariate.

```{r superpop}
markov_data <- sim_trajectories_brownian_gap(
  n_patients = N_PATIENTS,
  follow_up_time = FU,
  treatment_prob = 0.5,
  absorbing_state = 6, 
  seed = SEED,
  mu_treatment_effect = 0
)

# Add a baseline age covariate (one value per patient)
age_lookup <- tibble(
  id = sort(unique(markov_data$id)),
  age = rnorm(length(unique(markov_data$id)), mean = 65, sd = 12)
)
markov_data <- markov_data |>
  left_join(age_lookup, by = "id")

# Empirical SOPs (for reference)
plot_sops(markov_data, geom = "line", linetype_var = "tx", facet_var = NULL) + 
  scale_color_brewer(palette = "Dark2") +
  coord_cartesian(ylim = c(0, 1.0)) +
  labs(title = "Empirical State Occupancy Probabilities")
```

```{r}
# Calculate True Effect: Difference in time in state 1 (Home)
calc_time_in_state_diff(markov_data)
```

## Step 2: Prepare Markov Data

`prepare_markov_data()` creates the transition-level data used by the ordinal
Markov models. The models below use `rms::rcs(time, 4)` directly in the formula.
The fitted VGAM model stores the spline knot locations, and the SOP routines
reuse those locations when they build predictions over follow-up time.

```{r}
data <- prepare_markov_data(markov_data, absorbing_state = 6)
baseline_data <- data |> filter(time == 1)
```

## Step 3: Model Specification

We fit three proportional-odds models with increasing flexibility in how
transition probabilities can change over follow-up time. Each model keeps the
time spline inline in the formula, so the same model object can be used for
point estimates, nonparametric bootstrap inference, and simulation inference.

1.  **Model A (Simple PO):** a common spline-shaped time effect.

2.  **Model B (Treatment-time interaction):** allows the treatment effect to vary over time.

3.  **Model C (State-time interaction):** also allows the previous state's effect to vary over time.

### Fit the models

```{r}
m_simple <- vglm(
  ordered(y) ~ rcs(time, 4) + tx + age + yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE), 
  data = data
)

m_inter <- vglm(
  ordered(y) ~ rcs(time, 4) * tx + age + yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE), 
  data = data
)

m_state_time <- vglm(
  ordered(y) ~ rcs(time, 4) * tx + age + yprev + rcs(time, 4):yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE), 
  data = data
)
```

## Step 4: Bootstrap Marginalized SOPs

We compute the standardized (marginalized) State Occupancy Probabilities using bootstrapping. We standardize over the observed distribution of baseline covariates (like `yprev`) to get the population-average effect of treatment.

```{r}
run_boot <- function(model, name) {
  cat(paste0("Bootstrapping ", name, "...\n"))
  avg_sops(
    model, 
    newdata = data, 
    variables = list(tx = c(0, 1)),  # Compare treatment vs control
    times = 1:FU,
    ylevels = 1:6,
    absorb = "6",
    id_var = "id"
  ) |> 
     inferences(
      method = "bootstrap",
      n_sim = R, 
      workers = 6,
      conf_level = 0.95,
      return_draws = TRUE
    )
}

set.seed(SEED)
bs_simple <- run_boot(m_simple, "Model A")
bs_inter  <- run_boot(m_inter,  "Model B")
bs_state_time <- run_boot(m_state_time, "Model C")
```


### Simulation Inference: MVN vs Score Bootstrap

1. `engine = "mvn"`: coefficient draws from a multivariate normal approximation.
2. `engine = "score_bootstrap"`: one-step score perturbation with cluster-level
   exponential weights (implemented for `robcov_vglm` models).

```{r}
run_sim <- function(model, name, engine = "mvn") {
  cat(paste0("Simulation (", engine, ") - ", name, "...\n"))
  avg_sops(
    robcov_vglm(model, cluster = data$id), #model needs to have cluster robust SEs
    newdata = baseline_data, 
    variables = "tx",  # Compare treatment vs control
    times = 1:FU,
    ylevels = 1:6,
    absorb = "6",
    id_var = "id"
  ) |> 
     inferences(
      method = "simulation",
      engine = engine,
      n_sim = R, 
      conf_level = 0.95,
      workers = 1,
      return_draws = TRUE
    )
}

run_sim_set <- function(engine = "mvn") {
  list(
    simple = run_sim(
      m_simple,
      "Model A",
      engine = engine
    ),
    inter = run_sim(
      m_inter,
      "Model B",
      engine = engine
    ),
    state_time = run_sim(
      m_state_time,
      "Model C",
      engine = engine
    )
  )
}

set.seed(SEED)
t1 <- Sys.time()
res_mvn <- run_sim_set(engine = "mvn")
t2 <- Sys.time()

t2 - t1

set.seed(SEED)
res_score <- run_sim_set(engine = "score_bootstrap")

bs_simple_mvn <- res_mvn$simple
bs_inter_mvn  <- res_mvn$inter
bs_state_time_mvn <- res_mvn$state_time

bs_simple_score <- res_score$simple
bs_inter_score  <- res_score$inter
bs_state_time_score <- res_score$state_time
```

```{r}
# Compare average CI widths for model A: MVN vs score bootstrap
bind_rows(
  bs_simple_mvn |>
    mutate(setting = "MVN"),
  bs_simple_score |>
    mutate(setting = "Score bootstrap")
) |>
  group_by(setting, tx, state) |>
  summarise(mean_ci_width = mean(conf.high - conf.low), .groups = "drop") |>
  filter(state == 1) |>
  knitr::kable(digits = 3)
```

## Step 5: Visualization

### 1. State Occupancy Probabilities

We plot the SOPs with 95% confidence bands for all three models.

```{r}
plot_sop_estimates <- function(x, title) {
  ggplot(x, aes(x = time)) +
    geom_line(aes(y = estimate, linetype = factor(tx), group = interaction(state, tx))) +
    geom_ribbon(aes(
      ymin = conf.low,
      ymax = conf.high,
      fill = factor(state),
      group = interaction(state, tx)
    ), alpha = 0.2, color = NA) +
    scale_fill_brewer(palette = "Dark2") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = title)
}

p1 <- plot_sop_estimates(bs_simple, "Model A: PO")
p2 <- plot_sop_estimates(bs_inter, "Model B: PO + Interaction")
p3 <- plot_sop_estimates(bs_state_time, "Model C: PO + State-Time Interaction")

# Combine using patchwork
(p1 | p2 | p3) +
  plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "Nonparametric Bootstrap")
```

### 2. MVN and Score Bootstrap Simulation

```{r}
p1_mvn <- plot_sop_estimates(bs_simple_mvn, "Model A: PO")
p2_mvn <- plot_sop_estimates(bs_inter_mvn, "Model B: PO + Interaction")
p3_mvn <- plot_sop_estimates(bs_state_time_mvn, "Model C: PO + State-Time Interaction")

# Combine using patchwork
(p1_mvn | p2_mvn | p3_mvn) +
  plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "MVN Simulation")
```

```{r}
p1_score <- plot_sop_estimates(bs_simple_score, "Model A: PO")
p2_score <- plot_sop_estimates(bs_inter_score, "Model B: PO + Interaction")
p3_score <- plot_sop_estimates(bs_state_time_score, "Model C: PO + State-Time Interaction")

(p1_score | p2_score | p3_score) +
  plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "Score Bootstrap Simulation")
```

### 3. Difference in Time Spent in State 1 (Home)

We calculate the restricted mean time spent in State 1 (Home) over follow-up
and visualize the distribution of the difference (Treatment - Control).

```{r fig.height=6, fig.width=4, out.width="70%"}
# Helper to extract difference
get_diff_dist <- function(bs_obj, label) {
  bs_obj |> 
    get_draws() |> 
    group_by(tx, state, draw_id) |> 
    summarise(time_in_state = sum(draw), .groups = "drop") |>
    pivot_wider(id_cols = c("draw_id", "state"), names_from = tx, values_from = "time_in_state", names_prefix = "tx_") |> 
    mutate(delta = tx_1 - tx_0) |> 
    mutate(model = label)
}

# Combine results
diff_res <- bind_rows(
  get_diff_dist(bs_simple, "Model A (Simple)"),
  get_diff_dist(bs_inter,  "Model B (Interact)"),
  get_diff_dist(bs_state_time, "Model C (State-time)")
)

# Plot
plot_diff_boot <- ggplot(diff_res |> filter(state == 1), aes(x = delta, y = model, fill = model)) +
  stat_dots(side = "bottom", scale = 0.4, color = NA) +
  stat_halfeye(alpha = 0.6, scale = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Difference in Time Spent in State 1 (Home)",
    subtitle = "Bootstrap",
    x = "Difference (Days)",
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

plot_diff_boot
```

```{r}

diff_res_mvn <- bind_rows(
  get_diff_dist(bs_simple_mvn, "Model A (Simple)"),
  get_diff_dist(bs_inter_mvn,  "Model B (Interact)"),
  get_diff_dist(bs_state_time_mvn, "Model C (State-time)")
)

diff_res_score <- bind_rows(
  get_diff_dist(bs_simple_score, "Model A (Simple)"),
  get_diff_dist(bs_inter_score,  "Model B (Interact)"),
  get_diff_dist(bs_state_time_score, "Model C (State-time)")
)

# Plot
plot_diff_mvn <- ggplot(diff_res_mvn |> filter(state == 1), aes(x = delta, y = model, fill = model)) +
  stat_dots(side = "bottom", scale = 0.4, color = NA) +
  stat_halfeye(alpha = 0.6, scale = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Difference in Time Spent in State 1 (Home)",
    subtitle = "MVN",
    x = "Difference (Days)",
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

plot_diff_score <- ggplot(diff_res_score |> filter(state == 1), aes(x = delta, y = model, fill = model)) +
  stat_dots(side = "bottom", scale = 0.4, color = NA) +
  stat_halfeye(alpha = 0.6, scale = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Difference in Time Spent in State 1 (Home)",
    subtitle = "Score bootstrap",
    x = "Difference (Days)",
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(-2.5, 2.5))

plot_diff_boot / plot_diff_mvn / plot_diff_score
```

### Summary of Estimates

```{r}
bind_rows(
  diff_res |> mutate(setting = "Bootstrap"),
  diff_res_mvn |> mutate(setting = "MVN"),
  diff_res_score |> mutate(setting = "Score bootstrap")
) |>
  group_by(setting, model, state) |>
  summarise(
    mean_tx_1 = mean(tx_1),
    mean_tx_0 = mean(tx_0),
    Mean_Diff = mean(delta),
    Lower_2.5 = quantile(delta, 0.025),
    Upper_97.5 = quantile(delta, 0.975),
    CI_width = Upper_97.5 - Lower_2.5,
    .groups = "drop"
  ) |>
  arrange(model) |> 
  filter(state == 1) |> 
  knitr::kable(digits = 2)
```

