---
title: "Visit-Time SOPs with Irregular ACTT-2 Assessments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visit-Time SOPs with Irregular ACTT-2 Assessments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

This example starts from ACTT-2-like daily trajectories, keeps only four
assessment days, fits the Markov model on a factor-valued visit index, adds
simulation-based uncertainty, and then maps the SOPs back to elapsed days as a
separate post-processing step.

```{r}
library(markov.misc)
library(VGAM)
library(ggplot2)

set.seed(20260513)

visit_days <- c(3, 7, 14, 21)
visit_levels <- paste0("visit_", seq_along(visit_days))
state_levels <- as.character(1:8)
time_map <- stats::setNames(visit_days, visit_levels)
```

First simulate daily ACTT-2-style ordinal outcomes, then retain the scheduled
assessment days. The visit index is a factor; the real days are kept in a
separate column and are not used by the transition model.

```{r}
actt2_daily <- sim_actt2_brownian(n_patients = 250, seed = 20260513)

actt2_visits <- actt2_daily[actt2_daily$time %in% visit_days, ]
actt2_visits <- actt2_visits[order(actt2_visits$id, actt2_visits$time), ]
actt2_visits$day <- actt2_visits$time
actt2_visits$visit <- ave(actt2_visits$day, actt2_visits$id, FUN = seq_along)
actt2_visits$visit_label <- paste0("visit_", actt2_visits$visit)

actt2_visits$yprev_visit <- ave(
  actt2_visits$y,
  actt2_visits$id,
  FUN = function(z) c(NA, head(z, -1))
)
actt2_visits$yprev_visit[actt2_visits$visit == 1] <-
  actt2_visits$yprev[actt2_visits$visit == 1]

model_data <- actt2_visits
model_data$time <- factor(model_data$visit_label, levels = visit_levels)
model_data$y <- ordered(model_data$y, levels = state_levels)
model_data$yprev <- factor(model_data$yprev_visit, levels = state_levels)

model_data <- model_data[as.character(model_data$yprev) != "8", ]

head(model_data[c("id", "tx", "day", "time", "yprev", "y")])
```

The model sees `time` as the visit scale. Use `vglm.markov()` as the package
entrypoint for VGAM Markov workflows, then wrap the fit with `robcov_vglm()` so
simulation inference uses patient-clustered robust standard errors. With
`times = NULL`, `avg_sops()` resolves the prediction grid from the fitted factor
levels. The plain `avg_sops()` result contains the standardized point estimates.

```{r}
fit <- vglm.markov(
  ordered(y) ~ tx + time + yprev,
  family = VGAM::cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data
)
fit_robust <- robcov_vglm(fit, cluster = model_data$id)

baseline_data <- model_data[!duplicated(model_data$id), ]

avg_visit_sops <- avg_sops(
  fit_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = NULL,
  ylevels = factor(state_levels),
  absorb = "8",
  pvarname = "yprev"
)

attr(avg_visit_sops, "avg_args")$times
head(avg_visit_sops)
```

Add uncertainty with coefficient draws from the multivariate normal
approximation. The vignette uses a small number of simulations to keep runtime
short; increase `n_sim` for analysis. Use `return_draws = TRUE` before
interpolation so intervals on the real-day scale are computed from interpolated
draws rather than from interpolated interval endpoints.

```{r}
set.seed(20260513)

avg_visit_sops <- inferences(
  avg_visit_sops,
  method = "simulation",
  engine = "mvn",
  n_sim = 75,
  return_draws = TRUE
)

head(avg_visit_sops)
```

Use `interpolate_sops()` when you want summaries on elapsed days. Here day 0 is
used as an empirical baseline anchor, but the returned SOPs start at day 1. The
same baseline distribution is used for both treatment counterfactuals.

```{r}
real_day_sops <- interpolate_sops(
  avg_visit_sops,
  time_map = time_map,
  xout = 1:max(visit_days),
  origin_time = 0
)

label_treatment <- function(x) {
  x <- as.character(x)
  out <- ifelse(
    x %in% c("1", "Treatment"),
    "Treatment",
    ifelse(x %in% c("0", "Control"), "Control", NA_character_)
  )
  factor(out, levels = c("Control", "Treatment"))
}

real_day_sops$tx_label <- label_treatment(real_day_sops$tx)
real_day_draws <- attr(real_day_sops, "simulation_draws")
if (!is.null(real_day_draws)) {
  real_day_draws$tx_label <- label_treatment(real_day_draws$tx)
  attr(real_day_sops, "simulation_draws") <- real_day_draws
}

head(real_day_sops)
```

```{r}
plot_sops(
  real_day_sops,
  geom = "line",
  facet_var = "tx_label",
  ribbon_alpha = 0.16
) +
  labs(
    x = "Day",
    title = "Interpolated SOPs with MVN Simulation Intervals"
  )
```

Because `real_day_sops` has already been interpolated to a daily grid starting
at day 1, `time_in_state()` integrates from day 1 using trapezoidal AUC.

```{r}
state_1_auc <- time_in_state(real_day_sops, target_states = "1")

state_1_auc[c("tx_label", "total_time")]
```
