---
title:  "Time as a Factor with orm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time as a Factor with orm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

Outcome assessments may happen at scheduled visits rather than a regular daily
grid, without the possibility of accurately reconstructing daily states. This
vignette starts from simulated daily ACTT-2-style trajectories, keeps only days
3, 7, 14, and 21, and fits an `orm_markov()` transition model where `time` is
a visit factor.

## Visit Data

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 220
SEED <- 20260531
VISIT_DAYS <- stats::setNames(c(3, 7, 14, 21), as.character(1:4))
STATE_LEVELS <- as.character(1:8)
ABSORB <- 8L

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  mu_treatment_effect = -0.5,
  seed = SEED
)

baseline_state <- raw_data[raw_data$time == 1, c("id", "yprev")]
names(baseline_state)[2] <- "baseline_yprev"

visit_data <- raw_data[
  raw_data$time %in% unname(VISIT_DAYS),
  c("id", "tx", "time", "y", "yprev")
]
visit_data <- merge(
  visit_data,
  baseline_state,
  by = "id",
  all.x = TRUE,
  sort = FALSE
)
visit_data <- visit_data[order(visit_data$id, visit_data$time), ]
visit_data$visit_day <- visit_data$time
visit_data$time <- factor(
  match(visit_data$visit_day, unname(VISIT_DAYS)),
  levels = seq_along(VISIT_DAYS),
  labels = names(VISIT_DAYS)
)
visit_data$yprev <- ave(
  visit_data$y,
  visit_data$id,
  FUN = function(y) c(NA_integer_, y[-length(y)])
)
first_visit <- !duplicated(visit_data$id)
visit_data$yprev[first_visit] <- visit_data$baseline_yprev[first_visit]
visit_data$baseline_yprev <- NULL

model_data <- prepare_markov_data(
  visit_data,
  absorbing_state = ABSORB,
  factor_previous = TRUE
)

head(model_data)
```

## Fit And Predict On The Visit Scale

For factor-valued visits, supply the visit labels explicitly with `times`.
The returned SOPs are on the visit scale.

```{r fit-model}
dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_visit <- orm_markov(
  y ~ tx + time + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  penalty = 0.1,
  maxit = 100
)

sop_visit <- avg_sops(
  fit_visit,
  variables = list(tx = c(0, 1)),
  times = names(VISIT_DAYS),
  ylevels = fit_visit$yunique,
  absorb = as.character(ABSORB)
) |> 
  inferences(n_sim = 100)

attr(sop_visit, "avg_args")$times
```

## Interpolate To Calendar Days

The model is fit on visit labels. `interpolate_sops()` maps those labels to
nominal calendar days, adds an empirical day-0 baseline anchor, and linearly
interpolates probabilities between observed visits.

```{r plot-sops}
sop_days <- interpolate_sops(
  sop_visit,
  time_map = VISIT_DAYS,
  xout = 0:max(VISIT_DAYS),
  origin_time = 0
)

plot_sops(
  sop_days,
  time_var = "time",
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  scale_x_continuous(breaks = c(0, unname(VISIT_DAYS))) +
  labs(
    x = "Calendar day",
    y = "State occupancy probability"
  )
```

## Contrasts

```{r}
cmp_days <- avg_comparisons(
  fit_visit,
  variables = list(tx = c(0, 1)),
  metric = "time_in_state",
  states = NULL, 
  comparison = "difference",
  times = names(VISIT_DAYS),
  ylevels = fit_visit$yunique,
  absorb = as.character(ABSORB),
  time_map = VISIT_DAYS,
  xout = 0:max(VISIT_DAYS),
  origin_time = 0,
  time_unit = "days"
) |>
  inferences(n_sim = 100)

plot_comparisons(cmp_days) +
  labs(y = "Difference in time in state (days)")
```
