---
title: "Fractional Weighted Bootstrap for SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fractional Weighted Bootstrap for SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

This vignette shows the fractional weighted bootstrap (FWB) engine for marginal
state occupancy probabilities.

The ordinary refit bootstrap samples patients with replacement, refits the
transition model, and recomputes SOPs. The FWB keeps every patient in every
bootstrap replicate, draws one positive exponential weight per patient, refits
the model with those row-expanded weights, and uses the same patient weights
when marginalizing SOPs. This can be useful when ordinary bootstrap resamples
occasionally create sparse or degenerate refit samples.

The number of bootstrap replicates is deliberately tiny for vignette runtime.

## Fit The Model

```{r setup}
library(markov.misc)
library(ggplot2)
library(dplyr)
library(rms)

N_PATIENTS <- 160
FOLLOW_UP <- 10
N_BOOT <- 30
SEED <- 20260606
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
) |> 
  group_by(id) |> 
  mutate(initial_y = first(y)) |> 
  ungroup()

fit <- orm_markov(
  y ~ tx + rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)
```

## Define Marginal SOPs

Because `fit` was created with `orm_markov(id_var = "id")`, `sops()` can
use the stored longitudinal transition data for refits while automatically
extracting one prediction row per patient for the SOP target.

```{r avg-sops}
avg <- sops(
  fit,
  by = "initial_y",
  times = seq_len(FOLLOW_UP),
  ylevels = fit$yunique,
  absorb = ABSORB
)

head(avg)
```

## Fractional Bootstrap

Use `engine = "fwb"`for fractional weighted refits.

```{r run-bootstrap}
set.seed(SEED)
sop_fwb <- inferences(
  avg,
  method = "bootstrap",
  engine = "fwb",
  n_sim = N_BOOT,
  workers = 1,
  return_draws = TRUE
)
```

## Plot

```{r compare}
plot_sops(
  sop_fwb,
  facet_var = "initial_y"
)
```
