---
title: "Fractional Weighted Bootstrap for SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fractional Weighted Bootstrap for SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

This vignette shows the fractional weighted bootstrap (FWB) engine for marginal
state occupancy probabilities.

The ordinary refit bootstrap samples patients with replacement, refits the
transition model, and recomputes SOPs. The FWB keeps every patient in every
bootstrap replicate, draws one positive exponential weight per patient, refits
the model with those row-expanded weights, and uses the same patient weights
when marginalizing SOPs. This can be useful when ordinary bootstrap resamples
occasionally create sparse or degenerate refit samples.

The number of bootstrap replicates is deliberately tiny for vignette runtime.

## Fit The Model

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 160
FOLLOW_UP <- 10
N_BOOT <- 30
SEED <- 20260606
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)

fit <- orm(
  y ~ tx + rcs(time, 4) + yprev,
  data = model_data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

c(
  patients = length(unique(model_data$id)),
  transitions = nrow(model_data),
  states = length(STATE_LEVELS)
)
```

## Define Marginal SOPs

Refit bootstrap inference needs the full longitudinal transition data in
`newdata`, not just one baseline row per patient, because each bootstrap draw
refits the transition model.

```{r avg-sops}
avg <- avg_sops(
  fit,
  newdata = model_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit$yunique,
  absorb = ABSORB,
  id_var = "id"
)

head(avg)
```

## Run Standard And Fractional Bootstrap

Use `engine = "standard"` for ordinary patient resampling and `engine = "fwb"`
for fractional weighted refits. The default bootstrap engine is standard, so
the first call could omit `engine`.

```{r run-bootstrap}
set.seed(SEED)
sop_standard <- inferences(
  avg,
  method = "bootstrap",
  engine = "standard",
  n_sim = N_BOOT,
  workers = 1,
  return_draws = TRUE
)

set.seed(SEED)
sop_fwb <- inferences(
  avg,
  method = "bootstrap",
  engine = "fwb",
  n_sim = N_BOOT,
  workers = 1,
  return_draws = TRUE
)

data.frame(
  engine = c(attr(sop_standard, "engine"), attr(sop_fwb, "engine")),
  successful_draws = c(
    attr(sop_standard, "n_successful"),
    attr(sop_fwb, "n_successful")
  ),
  weight_type = c(NA, attr(sop_fwb, "fwb_weight_type")),
  weight_scale = c(NA, attr(sop_fwb, "fwb_weight_scale"))
)
```

## Compare

The point estimates are the same because both inference calls start from the
same fitted model and marginal SOP target. The interval estimates differ
because the bootstrap draws are generated differently.

```{r compare}
comparison <- rbind(
  data.frame(engine = "Standard bootstrap", as.data.frame(sop_standard)),
  data.frame(engine = "Fractional weighted bootstrap", as.data.frame(sop_fwb))
)

plot_sops(
  comparison,
  geom = "line",
  facet_var = c("engine", "tx")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

The FWB metadata records the implemented weighting scheme: exponential
patient-level weights normalized to mean 1 for model fitting. The marginal SOP
averaging step normalizes the same patient weights to sum 1 within each
bootstrap replicate.
