---
title: "Full Proportional-Odds SOPs with Markov Fit Wrappers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full Proportional-Odds SOPs with Markov Fit Wrappers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

We fit the same full proportional-odds first-order transition model using two
backends:

1. `vglm_markov()`, the package-aware VGAM wrapper.
2. `orm_markov()`, the package-aware rms wrapper.

The vignette uses 200 simulated ACTT-2-style patients on the simulator's native
8-state outcome scale and a shortened follow-up window to keep computation time
short.

## Data

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260527
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
```

## Recursion

Let $Y_{it}$ be the ordinal outcome for patient $i$ at prediction time
$t$, and let $\mathbf{X}_{it}$ collect treatment, time, and other covariates.
The state occupancy probability (SOP) for state $l$ is

$$
S_{it}(l) =
  \Pr(Y_{it} = l \mid \mathbf{X}_{i1}, \ldots, \mathbf{X}_{it}, Y_{i0}).
$$

For a first-order transition model, the conditional probability of moving from
state $k$ at time $t - 1$ to state $l$ at time $t$ is

$$
T_{it}(l \mid k) =
  \Pr(Y_{it} = l \mid Y_{i,t-1} = k, \mathbf{X}_{it}).
$$

For a single patient, collect these probabilities in the transition matrix
$\mathbf{P}_{it}$, where row $k$, column $l$ is
$P_{it,kl} = T_{it}(l \mid k)$. If
$\mathbf{s}_{it}$ is the row vector with entries $S_{it}(l)$, then the
matrix recursion is

$$
\mathbf{s}_{i0} = \mathbf{e}_{Y_{i0}}, \qquad
\mathbf{s}_{it} = \mathbf{s}_{i,t-1} \mathbf{P}_{it}, \quad t = 1,\ldots,d.
$$

The implementation batches this same calculation across patients. For each
possible previous state $k$, it predicts an $n \times K$ matrix
$\mathbf{P}^{(k)}_t$ whose $i,l$ entry is $T_{it}(l \mid k)$. The
batched update is

$$
S_{t,il} =
  \sum_{k \notin \mathcal{A}}
    S_{t-1,ik} P^{(k)}_{t,il}.
$$

Equivalently, in vector form over patients,

$$
\mathbf{S}_{t,l} =
  \sum_{k \notin \mathcal{A}}
    \mathbf{S}_{t-1,k} \odot \mathbf{P}^{(k)}_{t,l},
$$

where $\odot$ denotes elementwise multiplication and $\mathcal{A}$ is the set
of absorbing states. For an absorbing state $a \in \mathcal{A}$, the code
does not need a model prediction from $a$; it carries that probability mass
forward:

$$
\mathbf{S}_{t,a} \leftarrow
  \mathbf{S}_{t,a} + \mathbf{S}_{t-1,a}.
$$

The marginal arm-specific SOP used below is the average of these patient-level
SOPs after setting treatment to the desired counterfactual arm:

$$
\bar{S}_t(l; a) =
  n^{-1} \sum_{i=1}^n S_{it}(l; \operatorname{tx}=a).
$$

## Fit Models

```{r fit-models}
fit_vglm <- vglm_markov(
  ordered(y) ~ tx + rms::rcs(time, 4) + yprev,
  family = VGAM::cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_orm <- orm_markov(
  y ~ tx + rms::rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

data.frame(
  backend = c("VGAM vglm", "rms orm"),
  log_likelihood = c(
    as.numeric(stats::logLik(fit_vglm$vglm_fit)),
    as.numeric(stats::logLik(fit_orm))
  ),
  coefficients = c(length(stats::coef(fit_vglm)), length(stats::coef(fit_orm)))
)
```

## Compute And Plot SOPs

```{r compute-sops}
sop_vglm <- avg_sops(
  fit_vglm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = factor(STATE_LEVELS, levels = STATE_LEVELS),
  absorb = ABSORB
) |> 
  inferences(
    n_sim = 100
  )

sop_orm <- avg_sops(
  fit_orm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB
) |> 
  inferences(n_sim = 100)

combined_sops <- rbind(
  data.frame(backend = "VGAM vglm", as.data.frame(sop_vglm)),
  data.frame(backend = "rms orm", as.data.frame(sop_orm))
)

head(combined_sops)
```

```{r plot-sops}
plot_sops(
  combined_sops,
  geom = "line",
  facet_var = c("backend", "tx")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## References

Rohde MD, French B, Stewart TG, Harrell FE. Bayesian transition models for
ordinal longitudinal outcomes. *Statistics in Medicine*. 2024. doi:
10.1002/sim.10133.