---
title: "Many Outcome Levels with a Previous-State Spline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Many Outcome Levels with a Previous-State Spline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

When the outcome has many ordered levels, a categorical previous-state effect can become too large. We can instead model the previous state effect using a restricted cubic spline to allow for non-linear effects.

This vignette simulates a 30-level quality-of-life outcome and fits an
`orm_markov()` transition model with `rms::rcs(yprev, 5)`.

## Simulate Many Ordered States

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 220
FOLLOW_UP <- 14
N_QOL <- 30
STATE_LEVELS <- as.character(seq_len(N_QOL))
SEED <- 20260530

thresholds <- stats::qnorm(seq(0.02, 0.98, length.out = N_QOL - 1L))

raw_data <- sim_trajectories_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  n_states = length(STATE_LEVELS),
  thresholds = thresholds,
  allowed_start_state = as.integer(12:28),
  absorbing_state = NULL,
  x0_sd = 1.1,
  sigma_rw = 0.18,
  mu_drift = -0.08,
  mu_quad = 0.005,
  mu_treatment_effect = 0.08,
  treatment_prob = 0.5,
  drift_change_times = NULL,
  latent_dist = "normal",
  seed = SEED
)

plot_sops(raw_data, geom = "bar")
```

## Fit `rcs(yprev, 5)`

`prepare_markov_data(..., factor_previous = FALSE)` keeps `yprev` numeric so it
can enter the model as a smooth previous-state effect.

```{r fit-model}
model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = NULL,
  factor_previous = FALSE
)
model_data$y <- ordered(model_data$y, levels = STATE_LEVELS)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_many <- orm_markov(
  y ~ rms::rcs(time, 4) + tx + rms::rcs(yprev, 5),
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  penalty = 0.1,
  maxit = 100
)

fit_many
```

## Diagnostics

```{r}
ec <- plot_correlation(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y")

mc <- plot_correlation(
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)

ec
mc
```

```{r}
et <- plot_transitions(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y")

mt <- plot_transitions(
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)

et
mt

td <- plot_transitions(
  fit_many,
  comparison = "difference",
  variables = list(tx = c(0, 1)),
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)
td
```

```{r}
plot_variogram(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
  )

plot_variogram(
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)
```

## SOPs

```{r avg-sops}
sop_many <- avg_sops(
  fit_many,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  p_var = "yprev"
) |> 
  inferences(n_draws = 10)
```

With many states, a stacked distribution is usually easier to read than one
line per state.

```{r plot-sops}
plot_sops(
  sop_many,
  geom = "bar",
  facet_var = "tx",
  n_draws = 0
) +
  scale_x_continuous(breaks = c(1, 7, 14)) +
  labs(
    x = "Week",
    y = "State occupancy probability"
  )
```

## Constrasts

We can calcualte contrasts for time in state or weeks benefit.

```{r}
c <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)), 
  estimand = "time_in_state",
  state_sets = list(good = 1:10, average = 11:20, bad = 21:30),
) |> 
  inferences(
    n_draws = 10
  )

plot_comparisons(c)
```

```{r}
tb <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "time_benefit"
) |> 
  inferences(
    n_draws = 10
  )

plot_comparisons(tb)
```

```{r}
sc <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "sop"
) |> 
  inferences(
    n_draws = 10
  )

plot_comparisons(sc, facet_var = "state_set")
```
