---
title: "MVN Simulation Versus Bootstrap for orm() SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVN Simulation Versus Bootstrap for orm() SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

This vignette fits one full proportional-odds `rms::orm()` transition model and
compares two uncertainty routes for marginal SOPs:

1. Multivariate-normal (MVN) simulation from the fitted coefficient covariance.
2. Patient bootstrap with model refitting.

Simulation from MVN keeps the standardization sample fixed and varies the model coefficients. 
The nonparametric patient bootstrap resamples the empirical covariate and baseline-state distribution.

The number of draws is deliberately tiny for vignette runtime.

## Fit The Model

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 200
FOLLOW_UP <- 10
N_SIM <- 6
SEED <- 20260602
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

set.seed(SEED)
age_lookup <- data.frame(
  id = sort(unique(raw_data$id)),
  age = stats::rnorm(length(unique(raw_data$id)), mean = 65, sd = 12)
)
raw_data <- merge(raw_data, age_lookup, by = "id", sort = FALSE)
raw_data <- raw_data[order(raw_data$id, raw_data$time), ]

model_data <- prepare_markov_data(raw_data, absorbing_state = as.integer(ABSORB))
baseline_data <- model_data[!duplicated(model_data$id), , drop = FALSE]

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_orm <- rms::orm(
  y ~ tx + age + rms::rcs(time, 4) + yprev,
  data = model_data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

fit_robust <- rms::robcov(fit_orm, cluster = model_data$id)
robust_vcov <- as.matrix(fit_robust$var)
robust_vcov <- (robust_vcov + t(robust_vcov)) / 2
```

## Create SOP Targets

Bootstrap inference needs the full transition data so each bootstrap draw can
resample patients and refit the model. MVN simulation uses only the
baseline rows because the fitted model and covariance are already fixed.

```{r avg-sops}
avg_mvn <- avg_sops(
  fit_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB,
  id_var = "id"
)

avg_boot <- avg_sops(
  fit_orm,
  newdata = model_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB,
  id_var = "id"
)
```

## Run Both Methods

```{r run-inference}
set.seed(SEED)
sop_mvn <- inferences(
  avg_mvn,
  method = "simulation",
  engine = "mvn",
  n_sim = N_SIM,
  vcov = robust_vcov,
  workers = 1,
  return_draws = TRUE
)

set.seed(SEED)
sop_boot <- inferences(
  avg_boot,
  method = "bootstrap",
  n_sim = N_SIM,
  workers = 1,
  return_draws = TRUE
)

data.frame(
  method = c("MVN simulation", "Patient bootstrap"),
  successful_draws = c(
    attr(sop_mvn, "n_successful"),
    attr(sop_boot, "n_successful")
  )
)
```

## Compare

```{r compare}
comparison <- rbind(
  data.frame(method = "MVN simulation", as.data.frame(sop_mvn)),
  data.frame(method = "Patient bootstrap", as.data.frame(sop_boot))
)

plot_sops(
  comparison,
  geom = "line",
  facet_var = c("method", "tx")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

The point estimates are based on the same transition model. The confidence-band widths differ (especially close to baseline) because the two procedures target the variance of different estimands.

## References

Magirr D, Wang C, Przybylski A, Baillie M. Estimating the Variance of Covariate-Adjusted Estimators of Average Treatment Effects in Clinical Trials With Binary Endpoints. Pharm Stat. 2025 Jul-Aug;24(4):e70021. doi: 10.1002/pst.70021.
