---
title: "MVN Simulation Versus Bootstrap for orm_markov() SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVN Simulation Versus Bootstrap for orm_markov() SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

This vignette fits one full proportional-odds `orm_markov()` transition model and
compares two uncertainty routes for marginal SOPs:

1. Multivariate-normal (MVN) simulation from the fitted coefficient covariance.
2. Patient bootstrap with model refitting.

Simulation from MVN keeps the standardization sample fixed and varies the model coefficients. 
The nonparametric patient bootstrap resamples the empirical covariate and baseline-state distribution.

The number of draws is deliberately tiny for vignette runtime.

## Fit The Model

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 200
FOLLOW_UP <- 10
N_DRAWS <- 50
SEED <- 20260602
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

set.seed(SEED)
age_lookup <- data.frame(
  id = sort(unique(raw_data$id)),
  age = stats::rnorm(length(unique(raw_data$id)), mean = 65, sd = 12)
)
raw_data <- merge(raw_data, age_lookup, by = "id", sort = FALSE)
raw_data <- raw_data[order(raw_data$id, raw_data$time), ]

model_data <- prepare_markov_data(raw_data, absorbing_state = as.integer(ABSORB))

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_orm <- orm_markov(
  y ~ tx + age + rms::rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

robust_vcov <- as.matrix(fit_orm$var)
robust_vcov <- (robust_vcov + t(robust_vcov)) / 2
```

## Create SOP Targets

The wrapper stores the full transition data for bootstrap refits and lets
`avg_sops()` extract the baseline prediction rows automatically. MVN simulation
uses the cluster-robust covariance returned by `orm_markov(id_var = "id")`.

```{r avg-sops}
avg <- avg_sops(
  fit_orm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
)

# avg_boot <- avg_sops(
#   fit_orm,
#   variables = list(tx = c(0, 1)),
#   times = seq_len(FOLLOW_UP),
#   y_levels = fit_orm$yunique,
#   absorb = ABSORB
# )
```

## Run Both Methods

```{r run-inference}
set.seed(SEED)
sop_mvn <- inferences(
  avg,
  method = "mvn",
  n_draws = N_DRAWS,
  vcov = robust_vcov,
  workers = 1,
  return_draws = TRUE
)

set.seed(SEED)
sop_boot <- inferences(
  avg,
  method = "bootstrap",
  n_draws = N_DRAWS,
  workers = 1,
  return_draws = TRUE
)

data.frame(
  method = c("MVN simulation", "Patient bootstrap"),
  successful_draws = c(
    attr(sop_mvn, "n_successful"),
    attr(sop_boot, "n_successful")
  )
)
```

## Compare

```{r compare}
comparison <- rbind(
  data.frame(method = "MVN simulation", as.data.frame(sop_mvn)),
  data.frame(method = "Patient bootstrap", as.data.frame(sop_boot))
)

plot_sops(
  comparison,
  geom = "line",
  facet_var = c("method", "tx")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

The point estimates are based on the same transition model. The confidence-band widths differ (especially close to baseline) because the two procedures target the variance of different estimands.

## References

Magirr D, Wang C, Przybylski A, Baillie M. Estimating the Variance of Covariate-Adjusted Estimators of Average Treatment Effects in Clinical Trials With Binary Endpoints. Pharm Stat. 2025 Jul-Aug;24(4):e70021. doi: 10.1002/pst.70021.
