---
title: "Nonlinear Previous-State Effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonlinear Previous-State Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(markov.misc)
library(dplyr)
library(ggplot2)
library(VGAM)
library(rms)
```

The previous state can be modeled categorically, linearly, or with a nonlinear
numeric basis. The default remains categorical for backward compatibility:

```{r categorical-prep}
set.seed(20260505)

raw_data <- sim_trajectories_brownian(
  n_patients = 100,
  follow_up_time = 8,
  n_states = 10,
  thresholds = seq(-3, 3, length.out = 9),
  allowed_start_state = as.integer(2:9),
  absorbing_state = 10,
  x0_sd = 2,
  sigma_rw = 0.25,
  seed = 20260505
)

categorical_data <- prepare_markov_data(raw_data, absorbing_state = 10)
is.factor(categorical_data$yprev)
```

For high-cardinality ordinal states, a categorical previous-state effect can use
too many degrees of freedom. Set `factor_previous = FALSE` to keep `yprev`
numeric and use a linear or spline effect instead:

```{r numeric-prep}
numeric_data <- prepare_markov_data(
  raw_data,
  absorbing_state = 10,
  factor_previous = FALSE
)

is.numeric(numeric_data$yprev)
```

A linear previous-state effect uses one degree of freedom:

```{r linear-fit}
fit_linear <- vglm.markov(
  ordered(y) ~ rms::rcs(time, 4) + tx + yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = numeric_data
)
```

A restricted cubic spline allows nonlinear transition effects while still using
far fewer degrees of freedom than a full categorical previous-state model:

```{r spline-fit}
fit_spline <- vglm.markov(
  ordered(y) ~ rms::rcs(time, 4) + tx + rms::rcs(yprev, 6),
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = numeric_data
)
```

The same SOP workflow then applies. `vglm.markov()` stores the design-matrix
metadata needed for the fast simulation path when inline spline terms are used.
For inference, wrap the model in `robcov_vglm()` and pass the `avg_sops()`
result to `inferences()`:

```{r sops}
baseline <- numeric_data[numeric_data$time == 1, , drop = FALSE]
fit_spline_robust <- robcov_vglm(fit_spline, cluster = numeric_data$id)

avg <- avg_sops(
  fit_spline_robust,
  newdata = baseline,
  variables = list(tx = c(0, 1)),
  times = 1:8,
  ylevels = 1:10,
  absorb = "10",
  pvarname = "yprev",
  id_var = "id"
)

head(avg)
```

```{r sops-inference}
set.seed(20260506)

avg_mvn <- inferences(
  avg,
  method = "simulation",
  engine = "mvn",
  n_sim = 50,
  return_draws = FALSE
)

head(avg_mvn)
```

The model-based SOPs should be compared with empirical SOPs from the raw
longitudinal trajectories, before `prepare_markov_data()` creates transition
rows. This keeps absorbing states carried forward in the observed trajectory:

```{r stacked-sops, fig.width = 7, fig.height = 5}
empirical_sops <- raw_data |>
  filter(time %in% 1:8) |>
  count(tx, time, state = y, name = "n") |>
  group_by(tx, time) |>
  mutate(
    estimate = n / sum(n),
    source = "Empirical"
  ) |>
  ungroup() |>
  select(tx, time, state, estimate, source)

model_sops <- avg_mvn |>
  mutate(
    state = as.integer(as.character(state)),
    source = "Spline model"
  ) |>
  select(tx, time, state, estimate, source)

bind_rows(empirical_sops, model_sops) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = 1:10)
  ) |>
  ggplot(aes(x = time, y = estimate, fill = state)) +
  geom_col(width = 0.85) +
  facet_grid(tx ~ source) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    x = "Time",
    y = "State occupation probability",
    fill = "State"
  )
```

For uncertainty, plot selected state-specific trajectories with their simulated
confidence intervals and the matching empirical SOPs as points:

```{r inference-plot, fig.width = 7, fig.height = 5}
states_to_plot <- c(1, 5, 10)

empirical_selected <- empirical_sops |>
  filter(state %in% states_to_plot) |>
  mutate(state = factor(state, levels = states_to_plot))

avg_mvn |>
  mutate(state = as.integer(as.character(state))) |>
  filter(state %in% states_to_plot) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    state = factor(state, levels = states_to_plot)
  ) |>
  ggplot(aes(x = time, y = estimate, color = state, fill = state)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.7) +
  geom_point(
    data = empirical_selected |>
      mutate(
        tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
        state = factor(state, levels = states_to_plot)
      ),
    aes(x = time, y = estimate),
    shape = 21,
    color = "black",
    size = 1.8,
    inherit.aes = FALSE
  ) +
  facet_wrap(vars(tx)) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    x = "Time",
    y = "State occupation probability",
    color = "State",
    fill = "State"
  )
```

The same pattern scales to larger ordinal outcomes: keep `yprev` numeric with
`factor_previous = FALSE`, then include the desired numeric basis in the model
formula.
