---
title: "Quasi-Continuous Outcomes with a Previous-State Spline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quasi-Continuous Outcomes with a Previous-State Spline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dpi = 300
)
```

```{r setup, message = FALSE}
library(markov.misc)
library(dplyr)
library(ggplot2)
library(rms)

N_PATIENTS <- 500L
FOLLOW_UP <- 28L
QOL_LEVELS <- 1:100
DEATH_STATE <- 101L
STATE_LEVELS <- c(QOL_LEVELS, DEATH_STATE)
SEED <- 20260511L
```

Ordinal Markov models are not limited to a handful of clinical states. This
vignette uses a deliberately hypothetical 100-level quality-of-life state scale,
coded here so larger values are worse, and adds one absorbing death state as
state 101. Death is intentionally rare: about 1% of patients enter the absorbing
state over 28 discrete weeks.

The example has one purpose: show that the same Markov-model workflow can handle
a quasi-continuous ordinal outcome when the previous state is kept numeric and
modeled flexibly with `rms::rcs(yprev, 6)`.

## Simulate a Hypothetical 101-State Outcome

We use `sim_trajectories_brownian()` to generate plausible longitudinal movement
on a fine ordinal scale. The thresholds divide most of the latent normal range
into 100 living quality-of-life states, then place death in a far upper tail so
that the absorbing state is rare. Patients start in worse living states only,
then improve until approximately week 15 before deteriorating again. The
treatment effect is a small additional negative drift, which shifts patients
toward better living states under this coding.

```{r simulate}
qol_thresholds <- c(
  stats::qnorm(seq(0.005, 0.995, length.out = 99L)),
  5
)

raw_data <- sim_trajectories_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  n_states = length(STATE_LEVELS),
  thresholds = qol_thresholds,
  allowed_start_state = as.integer(45:100),
  absorbing_state = DEATH_STATE,
  x0_sd = 1.2,
  sigma_rw = 0.18,
  mu_drift = -0.09,
  mu_quad = 0.006,
  mu_treatment_effect = -0.01,
  treatment_prob = 0.5,
  drift_change_times = NULL,
  latent_dist = "normal",
  seed = SEED
)

raw_data |>
  summarise(
    patients = n_distinct(id),
    follow_up_weeks = max(time),
    start_min = min(yprev[time == 1]),
    start_max = max(yprev[time == 1]),
    mean_week_1 = mean(y[time == 1 & y != DEATH_STATE]),
    mean_week_15 = mean(y[time == 15 & y != DEATH_STATE]),
    mean_week_28 = mean(y[time == FOLLOW_UP & y != DEATH_STATE]),
    observed_outcome_levels = n_distinct(y),
    death_trajectories = n_distinct(id[y == DEATH_STATE]),
    death_risk = death_trajectories / patients
  )
```

## Fit a Fast `orm()` Markov Model

`prepare_markov_data()` removes transitions after death, but the incident death
rows remain in the response. Setting `factor_previous = FALSE` is the important
step for a high-cardinality previous state: it keeps `yprev` numeric so the
model can use a low-dimensional spline instead of 100 indicator terms.

```{r prepare-data}
model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = DEATH_STATE,
  factor_previous = FALSE
)
model_data$y <- ordered(model_data$y, levels = STATE_LEVELS)

data.frame(
  transition_rows = nrow(model_data),
  outcome_levels = length(levels(model_data$y)),
  previous_state_is_numeric = is.numeric(model_data$yprev)
)
```

The outcome has 101 ordered states, so the model has many intercepts while death
is deliberately sparse. The small ridge penalty and higher iteration cap make
this vignette fit stable; the Markov-model piece is still the ordinary
`rms::orm()` fast path with stored `x` and `y`.

```{r fit-model}
dd <- datadist(model_data)
options(datadist = "dd")

fit_orm <- orm(
  y ~ rms::rcs(time, 4) + tx + rms::rcs(yprev, 6),
  data = model_data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE,
  maxit = 100,
  penalty = 0.1
)

fit_orm_robust <- robcov(fit_orm, cluster = model_data$id)
robust_vcov <- as.matrix(fit_orm_robust$var)
```

## State Occupation Probabilities

For the standardization target, we use one observed row for each living previous
state. This keeps the vignette fast while still asking `avg_sops()` to propagate
probability over all 101 outcome states.

```{r avg-sops}
standardization_data <- model_data |>
  filter(yprev %in% QOL_LEVELS) |>
  arrange(yprev, id, time) |>
  distinct(yprev, .keep_all = TRUE)

qol_sops <- avg_sops(
  fit_orm_robust,
  newdata = standardization_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = as.character(STATE_LEVELS),
  absorb = as.character(DEATH_STATE),
  pvarname = "yprev",
  id_var = "id"
)

qol_sops |>
  group_by(tx, time) |>
  summarise(total_probability = sum(estimate), .groups = "drop") |>
  summarise(
    min_total_probability = min(total_probability),
    max_total_probability = max(total_probability)
  )
```

The same object can be passed to `inferences()`. We use only 10 multivariate
normal simulations so the vignette remains quick; a real analysis should use
many more draws.

```{r mvn-inference}
set.seed(SEED + 1L)

qol_sops_mvn <- inferences(
  qol_sops,
  method = "simulation",
  engine = "mvn",
  n_sim = 10,
  vcov = robust_vcov,
  return_draws = FALSE
)

qol_sops_mvn |>
  filter(
    time == FOLLOW_UP,
    state %in% c("25", "50", "75", as.character(DEATH_STATE))
  ) |>
  select(tx, time, state, estimate, conf.low, conf.high)
```

The raw simulated trajectories also give empirical SOPs. These are arm-specific
observed proportions, while the model-derived SOPs above are standardized over
the common previous-state distribution in `standardization_data`.

```{r empirical-sops}
empirical_sops <- raw_data |>
  count(tx, time, state = y, name = "n") |>
  group_by(tx, time) |>
  mutate(estimate = n / sum(n)) |>
  ungroup() |>
  select(tx, time, state, estimate)

state_fill_values <- c(
  stats::setNames(
    grDevices::hcl.colors(length(QOL_LEVELS), "Viridis"),
    as.character(QOL_LEVELS)
  ),
  stats::setNames("#2B2B2B", as.character(DEATH_STATE))
)

state_fill_scale <- function() {
  scale_fill_manual(
    values = state_fill_values,
    breaks = c("1", "25", "50", "75", "100", as.character(DEATH_STATE)),
    labels = c("1", "25", "50", "75", "100", "Death"),
    name = "State",
    drop = FALSE
  )
}
```

With 101 states, separate lines for each state are not helpful. A stacked bar
chart keeps the full state-occupation distribution visible at each time point.
Both panels use the same fill scale: living quality-of-life states are shown on
one gradient, and death is a separate dark level so it remains easy to spot.

```{r stacked-sops}
bind_rows(
  empirical_sops |>
    mutate(
      state = as.character(state),
      source = "Empirical"
    ),
  qol_sops_mvn |>
    select(tx, time, state, estimate) |>
    mutate(
      state = as.character(state),
      source = "Model derived"
    )
) |>
  mutate(
    tx = factor(tx, levels = c(0, 1), labels = c("Control", "Treatment")),
    source = factor(source, levels = c("Empirical", "Model derived")),
    state = factor(as.character(state), levels = as.character(STATE_LEVELS))
  ) |>
  ggplot(aes(x = time, y = estimate, group = state, fill = state)) +
  geom_col(width = 0.85) +
  facet_grid(tx ~ source) +
  scale_x_continuous(breaks = c(1, 7, 15, 21, 28)) +
  scale_y_continuous(
    labels = function(x) paste0(round(100 * x), "%"),
    expand = expansion(mult = c(0, 0.01))
  ) +
  state_fill_scale() +
  labs(
    x = "Week",
    y = "State occupation probability"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )
```

The key pattern is the same as for a smaller ordinal state space: fit an ordinal
transition model, identify the absorbing state, then let `avg_sops()` iterate the
transition probabilities forward. For quasi-continuous outcomes, the practical
change is to keep `yprev` numeric and use a compact basis such as
`rms::rcs(yprev, 6)`.
