---
title: "Uncertainty for Marginal SOPs with orm()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty for Marginal SOPs with orm()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  fig.align = "center",
  dpi = 300,
  fig.retina = 2
)
```

## Setup

```{r setup, message = FALSE, warning = FALSE}
library(markov.misc)
library(rms)
library(ggplot2)

N_PATIENTS <- 400
N_SIM <- 100
SEED <- 237567
FOLLOW_UP <- 30
STATE_LEVELS <- 1:6
ABSORB <- "6"
METHOD_LEVELS <- c("Bootstrap", "MVN simulation", "Score bootstrap")
```

`N_SIM` is intentionally small to keep vignette build time reasonable. For an
analysis, use a larger value such as `1000` or `2000`.

## Step 1: Simulate Trial Data

We simulate a discrete-time process using a latent Brownian motion model. We
also add a baseline age covariate. Age is not part of the outcome-generating
mechanism, but including it in the fitted model illustrates standardization over
an additional baseline covariate.

```{r superpop}
markov_data <- sim_trajectories_brownian_gap(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_prob = 0.5,
  absorbing_state = 6,
  seed = SEED,
  mu_treatment_effect = 0
)

set.seed(SEED + 1)
age_lookup <- data.frame(
  id = sort(unique(markov_data$id)),
  age = rnorm(length(unique(markov_data$id)), mean = 65, sd = 12)
)

markov_data <- merge(
  markov_data,
  age_lookup,
  by = "id",
  all.x = TRUE,
  sort = FALSE
)
markov_data <- markov_data[order(markov_data$id, markov_data$time), ]
rownames(markov_data) <- NULL

plot_sops(
  markov_data,
  geom = "line",
  linetype_var = "tx",
  facet_var = NULL
) +
  scale_color_brewer(palette = "Dark2") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Empirical State Occupancy Probabilities",
    x = "Day",
    y = "State occupancy probability"
  )
```

```{r empirical-time-in-state}
calc_time_in_state_diff(markov_data, target_state = 1)
```

## Step 2: Prepare Markov Data

`prepare_markov_data()` creates the transition-level data used by the ordinal
Markov model. The model uses `rms::orm()` only, with a full proportional-odds
specification and an inline restricted cubic spline for time.

```{r prepare-data}
data <- prepare_markov_data(markov_data, absorbing_state = 6)
baseline_data <- data[!duplicated(data$id), , drop = FALSE]

c(
  longitudinal_rows = nrow(data),
  baseline_patients = nrow(baseline_data)
)
```

## Step 3: Fit One Full Proportional-Odds Model

The `orm()` fit is created with `x = TRUE` and `y = TRUE`. Those stored design
objects are used by the fast SOP path and by score-bootstrap inference.

```{r fit-model}
dd <- datadist(data)
options(datadist = "dd")

fit_orm <- orm(
  y ~ rms::rcs(time, 4) + tx + age + yprev,
  data = data,
  x = TRUE,
  y = TRUE,
  opt_method = "LM",
  scale = TRUE
)

fit_orm_robust <- robcov(fit_orm, cluster = data$id)

c(
  log_likelihood = as.numeric(logLik(fit_orm)),
  max_deriv = unname(fit_orm$stats["Max Deriv"])
)
```

## Step 4: Compute Marginal SOP Uncertainty

The same fitted model is used in three ways:

1. Nonparametric bootstrap: resample patients, refit the `orm()` model, and
   recompute marginalized SOPs.
2. MVN simulation: draw coefficients from a multivariate normal approximation
   using the robust covariance matrix.
3. Score bootstrap: use one-step score perturbations with patient-level
   exponential weights.

Bootstrap inference needs the full longitudinal transition data because it
refits the model. MVN and score-bootstrap inference only need baseline rows
for the SOP standardization target.

```{r avg-sops-inputs}
avg_for_bootstrap <- avg_sops(
  fit_orm,
  newdata = data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB,
  id_var = "id"
)

avg_for_simulation <- avg_sops(
  fit_orm_robust,
  newdata = baseline_data,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_orm$yunique,
  absorb = ABSORB,
  id_var = "id"
)
```

```{r run-inference}
set.seed(SEED)
time_bootstrap <- system.time({
  sops_bootstrap <- inferences(
    avg_for_bootstrap,
    method = "bootstrap",
    n_sim = N_SIM,
    workers = 1,
    conf_level = 0.95,
    return_draws = TRUE
  )
})

set.seed(SEED)
time_mvn <- system.time({
  sops_mvn <- inferences(
    avg_for_simulation,
    method = "simulation",
    engine = "mvn",
    n_sim = N_SIM,
    workers = 1,
    conf_level = 0.95,
    return_draws = TRUE
  )
})

set.seed(SEED)
time_score <- system.time({
  sops_score <- inferences(
    avg_for_simulation,
    method = "simulation",
    engine = "score_bootstrap",
    cluster = data$id,
    n_sim = N_SIM,
    workers = 1,
    conf_level = 0.95,
    return_draws = TRUE
  )
})
```

The timing table measures only the uncertainty step after the initial model fit
and point-estimate SOP objects have been created.

```{r timing-table}
timing <- data.frame(
  method = METHOD_LEVELS,
  elapsed_seconds = c(
    unname(time_bootstrap[["elapsed"]]),
    unname(time_mvn[["elapsed"]]),
    unname(time_score[["elapsed"]])
  ),
  successful_draws = c(
    attr(sops_bootstrap, "n_successful"),
    attr(sops_mvn, "n_successful"),
    attr(sops_score, "n_successful")
  )
)

knitr::kable(timing, digits = 2)
```

## Step 5: Compare SOP Estimates

```{r combine-sops}
inference_sops <- rbind(
  data.frame(method = "Bootstrap", as.data.frame(sops_bootstrap)),
  data.frame(method = "MVN simulation", as.data.frame(sops_mvn)),
  data.frame(method = "Score bootstrap", as.data.frame(sops_score))
)

inference_sops$method <- factor(
  inference_sops$method,
  levels = METHOD_LEVELS
)
inference_sops$tx <- factor(
  as.integer(as.character(inference_sops$tx)),
  levels = c(0, 1),
  labels = c("Control", "Treatment")
)
inference_sops$state <- factor(
  as.character(inference_sops$state),
  levels = as.character(STATE_LEVELS)
)
inference_sops$ci_width <- inference_sops$conf.high - inference_sops$conf.low

head(inference_sops)
```

```{r sop-plot}
ggplot(
  inference_sops,
  aes(x = time, y = estimate, color = state, fill = state)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high, group = state),
    alpha = 0.12,
    color = NA
  ) +
  geom_line(aes(group = state), linewidth = 0.6) +
  facet_grid(tx ~ method) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Marginal State Occupancy Probabilities",
    subtitle = "One full proportional-odds orm() model, three uncertainty estimators",
    x = "Day",
    y = "State occupancy probability",
    color = "State",
    fill = "State"
  )
```

```{r ci-width-summary}
ci_width_state1 <- inference_sops[
  inference_sops$state == "1",
  c("method", "tx", "ci_width")
]
ci_width_summary <- aggregate(
  ci_width ~ method + tx,
  data = ci_width_state1,
  FUN = mean
)

knitr::kable(ci_width_summary, digits = 3)
```

## Step 6: Compare State 1 Treatment Differences

For each draw, we sum the State 1 SOP over follow-up and subtract control from
treatment. This is the model-standardized treatment difference in expected days
spent in State 1.

```{r draw-differences}
draws_bootstrap <- get_draws(sops_bootstrap)
draws_bootstrap$method <- "Bootstrap"
draws_bootstrap <- draws_bootstrap[
  ,
  c("method", "draw_id", "tx", "state", "time", "draw")
]

draws_mvn <- get_draws(sops_mvn)
draws_mvn$method <- "MVN simulation"
draws_mvn <- draws_mvn[
  ,
  c("method", "draw_id", "tx", "state", "time", "draw")
]

draws_score <- get_draws(sops_score)
draws_score$method <- "Score bootstrap"
draws_score <- draws_score[
  ,
  c("method", "draw_id", "tx", "state", "time", "draw")
]

draws_all <- rbind(draws_bootstrap, draws_mvn, draws_score)
draws_all$method <- factor(draws_all$method, levels = METHOD_LEVELS)
draws_all$tx <- as.character(draws_all$tx)
draws_all$state <- as.character(draws_all$state)

draws_state1 <- draws_all[draws_all$state == "1", ]

time_in_state <- aggregate(
  draw ~ method + draw_id + tx,
  data = draws_state1,
  FUN = sum
)
names(time_in_state)[names(time_in_state) == "draw"] <- "time_in_state"
time_in_state$tx <- as.character(time_in_state$tx)

time_control <- time_in_state[
  time_in_state$tx == "0",
  c("method", "draw_id", "time_in_state")
]
names(time_control)[names(time_control) == "time_in_state"] <- "tx_0"

time_treatment <- time_in_state[
  time_in_state$tx == "1",
  c("method", "draw_id", "time_in_state")
]
names(time_treatment)[names(time_treatment) == "time_in_state"] <- "tx_1"

diff_draws <- merge(
  time_treatment,
  time_control,
  by = c("method", "draw_id"),
  all = FALSE
)
diff_draws$delta <- diff_draws$tx_1 - diff_draws$tx_0
diff_draws$method <- factor(as.character(diff_draws$method), levels = METHOD_LEVELS)

head(diff_draws)
```

```{r difference-plot, fig.height = 5, fig.width = 7}
ggplot(diff_draws, aes(x = delta, y = method, fill = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_violin(trim = FALSE, alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.8) +
  geom_point(
    position = position_jitter(width = 0, height = 0.05),
    alpha = 0.45,
    size = 1
  ) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Difference In Time Spent In State 1",
    subtitle = "Treatment minus control, by uncertainty estimator",
    x = "Difference in expected days",
    y = NULL,
    fill = "Method"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r difference-summary}
diff_summary <- data.frame(
  method = METHOD_LEVELS,
  mean_days = NA_real_,
  lower_2.5 = NA_real_,
  upper_97.5 = NA_real_,
  ci_width = NA_real_
)

for (i in seq_along(METHOD_LEVELS)) {
  x <- diff_draws$delta[as.character(diff_draws$method) == METHOD_LEVELS[i]]
  diff_summary$mean_days[i] <- mean(x)
  diff_summary$lower_2.5[i] <- unname(quantile(x, 0.025))
  diff_summary$upper_97.5[i] <- unname(quantile(x, 0.975))
  diff_summary$ci_width[i] <- diff_summary$upper_97.5[i] -
    diff_summary$lower_2.5[i]
}

knitr::kable(diff_summary, digits = 2)
```
