---
title: "Partial Proportional-Odds SOPs with vglm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Proportional-Odds SOPs with vglm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

Partial proportional-odds models can relax the proportional-odds assumption for selected predictors while keeping the rest of the transition model parsimonious. This vignette fits such a model with `vglm_markov()`, allowing the linear time basis to deviate from
proportional odds while treatment and previous state remain proportional.

## Simulate Data

We collapse the 8 state data into 6 states the make the linear deviation from proportional odds more stable.

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260528
STATE_LEVELS <- as.character(1:6)
ABSORB <- "6"

collapse_actt2_states <- function(data) {
  data$y <- ifelse(
    data$y %in% c(1L, 2L),
    1L,
    ifelse(data$y %in% c(3L, 4L), 2L, data$y - 2L)
  )
  data$yprev <- ifelse(
    data$yprev %in% c(1L, 2L),
    1L,
    ifelse(data$yprev %in% c(3L, 4L), 2L, data$yprev - 2L)
  )
  data
}

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
) |>
  collapse_actt2_states()

model_data <- prepare_markov_data(raw_data, absorbing_state = 6)
```

## Constraint
```{r fit-ppo}
form <- ordered(y) ~ tx + rms::rcs(time, 4) + yprev

fit_po <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)

constraints <- fit_po$vglm_fit@constraints
linear_time_term <- grep(
  "^rms::rcs\\(time, 4\\).*time$",
  names(constraints),
  value = TRUE
)[[1]]
n_thresholds <- nrow(constraints[[linear_time_term]])

yprev_terms <- grep("^yprev", names(constraints), value = TRUE)
constraints[yprev_terms] <- NULL
constraints[["yprev"]] <- cbind(PO_effect = rep(1, n_thresholds))
constraints[[linear_time_term]] <- cbind(
  PO_effect = rep(1, n_thresholds),
  linear_deviation = seq_len(n_thresholds) - 1
)

fit_ppo <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = FALSE),
  data = model_data,
  id_var = "id",
  constraints = constraints,
  etastart = fit_po$vglm_fit@predictors
)
fit_ppo$vglm_fit@call$constraints <- constraints

data.frame(
  model = c("Full PO", "Linear-time PPO"),
  log_likelihood = c(
    as.numeric(stats::logLik(fit_po$vglm_fit)),
    as.numeric(stats::logLik(fit_ppo$vglm_fit))
  ),
  coefficients = c(length(stats::coef(fit_po)), length(stats::coef(fit_ppo)))
)
```

## SOPs

```{r avg-sops}
sop_ppo <- avg_sops(
  fit_ppo,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = factor(STATE_LEVELS, levels = STATE_LEVELS),
  absorb = ABSORB
) |> 
  inferences(n_sim = 100)
```

```{r plot-ppo}
plot_sops(
  sop_ppo,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Model Diagnostics

These diagnostics compare empirical trajectory features with corresponding
features implied by the partial proportional-odds transition model.

```{r diagnostics-transitions-observed}
diagnostic_times <- 1:FOLLOW_UP

plot_transitions(
  raw_data,
  times = diagnostic_times,
  ylevels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model}
plot_transitions(
  fit_ppo,
  times = diagnostic_times,
  ylevels = STATE_LEVELS,
  absorb = ABSORB,
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed}
plot_correlation(
  raw_data,
  times = seq_len(FOLLOW_UP),
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model}
plot_correlation(
  fit_ppo,
  times = seq_len(FOLLOW_UP),
  ylevels = STATE_LEVELS,
  absorb = ABSORB,
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed}
plot_variogram(
  raw_data,
  times = seq_len(FOLLOW_UP)
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model}
plot_variogram(
  fit_ppo,
  times = seq_len(FOLLOW_UP),
  ylevels = STATE_LEVELS,
  absorb = ABSORB
) +
  labs(title = "Model-based correlation variogram")
```

## Contrast


```{r}
c <- avg_comparisons(
  fit_ppo,
  metric = "time_in_state",
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = factor(STATE_LEVELS, levels = STATE_LEVELS),
  absorb = ABSORB) |> 
  inferences(n_sim = 100)

plot_comparisons(c)
```
