---
title: "Mathematical Derivation: One-Step Score Bootstrap for Markov SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Derivation: One-Step Score Bootstrap for Markov SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 300
)
```

## Goal

This vignette derives the one-step Newton-Raphson approximation used in the
score bootstrap to efficiently estimate uncertainty in State Occupation
Probabilities (SOPs).

To build intuition before the full matrix derivation, we first work through a
single-intercept model where the geometry can be seen directly.

## Setup

```{r setup, message=FALSE}
library(markov.misc)
library(VGAM)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)

SEED <- 20260216
N <- 100
P_TRUE <- 0.6

set.seed(SEED)
dat <- tibble(
  id = seq_len(N),
  y_bin = rbinom(N, size = 1, prob = P_TRUE),
  y = ordered(y_bin, levels = c(0, 1))
)

fit <- vglm(
  y ~ 1,
  family = cumulative(reverse = TRUE),
  data = dat
)

# One-observation-per-cluster to mirror the score-bootstrap machinery
# while intentionally ignoring clustering structure.
fit_rb <- robcov_vglm(fit, cluster = dat$id)

beta_hat <- as.numeric(stats::coef(fit)[1])
scores <- as.numeric(fit_rb$scores[, 1])
bread <- as.numeric(fit_rb$bread[1, 1])
n_obs <- fit_rb$n
v_hat <- bread / n_obs

stopifnot(length(stats::coef(fit)) == 1)
stopifnot(abs(sum(scores)) < 1e-8)
stopifnot(isTRUE(all.equal(v_hat, as.numeric(stats::vcov(fit)[1, 1]), tolerance = 1e-10)))

y_num <- dat$y_bin
```

For this fixed dataset, the fitted intercept corresponds to:

```{r}
tibble(
  quantity = c("beta_hat", "plogis(beta_hat)", "sample mean(y)"),
  value = c(beta_hat, plogis(beta_hat), mean(y_num))
)
```

## Log-Likelihood Geometry in 1D

For Bernoulli outcomes with one intercept:

- log-likelihood: $\ell(\beta) = \sum_i \{y_i \beta - \log(1 + e^\beta)\}$,
- score: $U(\beta) = \sum_i (y_i - \text{logit}^{-1}(\beta))$,
- Hessian: $H(\beta) = -n\,p(\beta)\{1-p(\beta)\}$.

```{r ll-score-hess}
ll_fn <- function(beta, y) {
  sum(y * beta - log1p(exp(beta)))
}

score_fn <- function(beta, y) {
  sum(y - plogis(beta))
}

hess_fn <- function(beta, n) {
  p <- plogis(beta)
  -n * p * (1 - p)
}

beta_grid <- seq(beta_hat - 4, beta_hat + 4, length.out = 500)
geom_df <- tibble(
  beta = beta_grid,
  loglik = vapply(beta_grid, ll_fn, numeric(1), y = y_num),
  score = vapply(beta_grid, score_fn, numeric(1), y = y_num),
  hess = vapply(beta_grid, hess_fn, numeric(1), n = n_obs)
)

p_ll <- ggplot(geom_df, aes(beta, loglik)) +
  geom_line(linewidth = 0.8, color = "#2C3E50") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "Log-likelihood",
    x = expression(beta),
    y = expression(ell(beta))
  ) +
  theme_minimal()

p_score <- ggplot(geom_df, aes(beta, score)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.8, color = "#2874A6") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "First derivative (Score)",
    x = expression(beta),
    y = expression(U(beta))
  ) +
  theme_minimal()

p_hess <- ggplot(geom_df, aes(beta, hess)) +
  geom_line(linewidth = 0.8, color = "#117864") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "Second derivative (Hessian)",
    x = expression(beta),
    y = expression(H(beta))
  ) +
  theme_minimal()

(p_ll / p_score / p_hess) +
  plot_annotation(title = "Likelihood Geometry for a Single Intercept")
```

At the MLE (`beta_hat`), the score is approximately zero.

## Score Contributions and Exponential Reweighting

Keeping this dataset fixed, we now perturb the score equation the same way the
score bootstrap does.

`robcov_vglm()` gives observation-level score contributions $s_i$ at
$\hat\beta$. In this one-parameter case, these are scalar values.

For one score-bootstrap draw:

- sample weights $w_i \sim \text{Exp}(1)$,
- center them as $u_i = w_i - 1$,
- compute perturbed score at the MLE:
  $$
  U^*(\hat\beta) = \sum_i u_i s_i.
  $$

```{r single-perturbation}
set.seed(SEED + 1)
w_i <- rexp(n_obs, rate = 1)
u_i <- w_i - 1

contrib_df <- tibble(
  obs = seq_len(n_obs),
  score_i = scores,
  centered_weight = u_i,
  weighted_score_i = u_i * score_i
)

score_totals <- tibble(
  type = c("Original total score", "Perturbed total score"),
  value = c(sum(contrib_df$score_i), sum(contrib_df$weighted_score_i))
)

knitr::kable(score_totals, digits = 4)
```

```{r contribution-viz}
p_contrib <- ggplot(contrib_df, aes(obs, weighted_score_i)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_linerange(aes(ymin = 0, ymax = weighted_score_i), color = "#2874A6", alpha = 0.6) +
  labs(
    title = "Perturbed contribution per observation",
    x = "Observation",
    y = expression((w[i] - 1) * s[i])
  ) +
  theme_minimal()

p_contrib
```

## How the Total Score Curve Changes

The key geometric effect is that exponential reweighting tilts the score curve
and moves its root.

The unweighted score curve is:
$$
U(\beta) = \sum_i (y_i - \text{logit}^{-1}(\beta)).
$$

With one exponential draw, the weighted score curve is:
$$
U_w(\beta) = \sum_i w_i (y_i - \text{logit}^{-1}(\beta)).
$$

At $\hat\beta$, $U(\hat\beta)=0$, while generally $U_w(\hat\beta)\neq 0$.

```{r score-shift-viz}
score_curve_df <- tibble(
  beta = beta_grid,
  unweighted = vapply(beta_grid, score_fn, numeric(1), y = y_num),
  weighted = vapply(
    beta_grid,
    function(b) sum(w_i * (y_num - plogis(b))),
    numeric(1)
  )
) |>
  pivot_longer(cols = c(unweighted, weighted), names_to = "curve", values_to = "score")

overview <- ggplot(score_curve_df, aes(beta, score, color = curve)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "grey35") +
  scale_color_manual(
    values = c(unweighted = "#2C3E50", weighted = "#C0392B"),
    labels = c(unweighted = "U(beta)", weighted = "U_w(beta)")
  ) +
  labs(
    title = "Score curve before and after reweighting",
    x = expression(beta),
    y = "Score",
    color = NULL
  ) +
  theme_minimal()

zoomed <- ggplot(score_curve_df, aes(beta, score, color = curve)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "grey35") +
  scale_color_manual(
    values = c(unweighted = "#2C3E50", weighted = "#C0392B"),
    labels = c(unweighted = "U(beta)", weighted = "U_w(beta)")
  ) +
  coord_cartesian(xlim = c(0.5, 1.25), ylim = c(-5, 5)) +
  labs(
    title = "Score curve before and after reweighting (zoomed)",
    x = expression(beta),
    y = "Score",
    color = NULL
  ) +
  theme_minimal()

overview / zoomed
```

## One-Step Update as a Tangent Approximation

Given that shifted score curve, the package uses a one-step Newton-style update
to approximate the new root.

In 1D, the one-step score-bootstrap update used in the package is:
$$
\beta^* \approx \hat\beta + V(\hat\beta) \, U^*(\hat\beta),
$$

where $V(\hat\beta)=(-H(\hat\beta))^{-1}$, and in package terms
$V(\hat\beta) = \texttt{bread}/n$.

```{r tangent-viz}
U_star <- sum(contrib_df$weighted_score_i)
H_hat <- hess_fn(beta_hat, n_obs)
beta_one_step <- beta_hat + v_hat * U_star

weighted_score_fn <- function(beta) {
  sum(w_i * (y_num - plogis(beta)))
}

find_root <- function(f, center, step = 2, max_iter = 25) {
  left <- center - step
  right <- center + step
  f_left <- f(left)
  f_right <- f(right)
  iter <- 0

  while (sign(f_left) == sign(f_right) && iter < max_iter) {
    step <- step * 1.5
    left <- center - step
    right <- center + step
    f_left <- f(left)
    f_right <- f(right)
    iter <- iter + 1
  }

  if (sign(f_left) == sign(f_right)) {
    return(NA_real_)
  }

  stats::uniroot(f, c(left, right))$root
}

beta_exact <- find_root(weighted_score_fn, center = beta_hat)
stopifnot(is.finite(beta_exact))

tangent_df <- tibble(
  beta = beta_grid,
  weighted_score = vapply(beta_grid, weighted_score_fn, numeric(1)),
  tangent = U_star + H_hat * (beta - beta_hat)
)

ggplot(tangent_df, aes(beta)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(aes(y = weighted_score, color = "Weighted score curve"), linewidth = 1) +
  geom_line(aes(y = tangent, color = "Tangent at beta_hat"), linewidth = 1, linetype = "dotted") +
  geom_vline(aes(xintercept = beta_hat, color = "beta_hat"), linetype = "dotdash") +
  geom_vline(aes(xintercept = beta_one_step, color = "One-step beta*"), linetype = "dotdash") +
  geom_vline(aes(xintercept = beta_exact, color = "Exact weighted-MLE root"), linetype = "dotdash") +
  scale_color_manual(
    values = c(
      "Weighted score curve" = "#2C3E50",
      "Tangent at beta_hat" = "#7D3C98",
      "beta_hat" = "#1E8449",
      "One-step beta*" = "#D35400",
      "Exact weighted-MLE root" = "#C0392B"
    )
  ) +
  coord_cartesian(xlim = c(0.7, 1.0), ylim = c(-5, 5)) +
  labs(
    title = "One-step Newton update from the tangent line",
    x = expression(beta),
    y = "Weighted score",
    color = NULL
  ) +
  theme_minimal()
```

```{r one-step-comparison}
tibble(
  quantity = c("beta_hat", "U_star(beta_hat)", "beta_one_step", "beta_exact"),
  value = c(beta_hat, U_star, beta_one_step, beta_exact)
) |>
  knitr::kable(digits = 5)
```

------------------------------------------------------------------------

Now we translate that 1D picture into the full clustered, multi-parameter
setting used for Markov SOP inference.

In the figures above, we saw three key ingredients: a score root at the MLE,
exponential perturbation of the score equation, and a tangent-based one-step
root update. The derivation below is exactly the same logic, written in vector
and matrix form.

## 1. The Maximum Likelihood Framework

Let $\ell(\beta)$ be the log-likelihood of our model given parameters $\beta$.
In a clustered dataset with $G$ clusters (patients), the total log-likelihood
is the sum of the cluster-level log-likelihoods:
$$
\ell(\beta) = \sum_{g=1}^G \ell_g(\beta)
$$

The **Score vector** $U(\beta)$ is the gradient of the log-likelihood:
$$
U(\beta) = \nabla \ell(\beta) = \sum_{g=1}^G U_g(\beta), \quad
\text{where } U_g(\beta) = \frac{\partial \ell_g}{\partial \beta}
$$

``` r
# Cluster-level scores (aggregated from observation-level scores)
scores_clustered <- rowsum(scores, cluster_fac, reorder = FALSE)
```

The **Hessian matrix** $H(\beta)$ is the second derivative (curvature):
$$
H(\beta) = \nabla^2 \ell(\beta) = \sum_{g=1}^G H_g(\beta)
$$

At the Maximum Likelihood Estimate (MLE) $\hat{\beta}$, the total score is
exactly zero:
$$
U(\hat{\beta}) = \sum_{g=1}^G U_g(\hat{\beta}) = \mathbf{0}
$$

------------------------------------------------------------------------

## 2. The Perturbed Likelihood

In the intercept plots above, one exponential draw tilted the score curve and
moved its root. In the general setting, we do that at the cluster level.

In the score bootstrap, we assign a random weight $w_g \sim \text{Exp}(1)$ to
each cluster.

``` r
w_cluster <- stats::rexp(n_clusters, rate = 1)
```

We want to find the new parameter estimate $\beta^*$ that maximizes the
**weighted log-likelihood**:
$$
\ell^*(\beta) = \sum_{g=1}^G w_g \ell_g(\beta)
$$

The corresponding perturbed score function is:
$$
U^*(\beta) = \sum_{g=1}^G w_g U_g(\beta)
$$

We are looking for $\beta^*$ such that the perturbed score is zero:
$U^*(\beta^*) = \mathbf{0}$.

------------------------------------------------------------------------

## 3. One-Step Update in the Multi-Parameter Case

The tangent-line figure in 1D is the scalar version of the update below.

### From score root-finding to one-step update

In maximum likelihood estimation, we want to solve the perturbed score equation
$U^*(\beta^*) = \mathbf{0}$.

Start from the classical scalar Newton step:
$$
x_{new} = x_{old} - \frac{f(x_{old})}{f'(x_{old})}
$$

In multiple dimensions, the derivative is a matrix, so we use an inverse:
$$
x_{new} = x_{old} - [f'(x_{old})]^{-1} f(x_{old})
$$

Now substitute each symbol for our score-bootstrap setting:

| Standard Newton-Raphson         | Score Bootstrap Setting              |
|---------------------------------|--------------------------------------|
| $x$ = unknown                   | $\beta$ = model parameters           |
| $f(x)$ = function to zero out   | $U^*(\beta)$ = perturbed score       |
| $f'(x)$ = derivative            | $H(\beta)$ = Hessian                 |
| $x_{old}$ = current guess       | $\hat{\beta}$ = original MLE         |
| $x_{new}$ = updated value       | $\beta^*$ = perturbed parameter draw |

Taking one Newton step from $\hat{\beta}$ then gives:
$$
\beta^* \approx \hat{\beta} - [H(\hat{\beta})]^{-1} \cdot U^*(\hat{\beta})
$$

Because $H(\hat{\beta})$ is negative definite at a maximum, this is typically
written as:
$$
\beta^* \approx \hat{\beta} + [-H(\hat{\beta})]^{-1} \cdot U^*(\hat{\beta})
$$

In multiple dimensions, this uses a matrix inverse rather than scalar division.
A first-order Taylor justification is provided in the Appendix.

### Centered exponential multipliers

We simplify $U^*(\hat{\beta})$ using the fact that the original total score is
zero:
$$
U^*(\hat{\beta}) = \sum w_g U_g(\hat{\beta})
= \sum (w_g - 1) U_g(\hat{\beta}) + \underbrace{\sum U_g(\hat{\beta})}_{\mathbf{0}}
$$
Let $u_g = (w_g - 1)$ be the centered multipliers.

``` r
u_cluster <- w_cluster - 1
```

Then:
$$
U^* = \sum_{g=1}^G u_g U_g(\hat{\beta})
$$

``` r
U_star <- as.vector(crossprod(u_cluster, scores_clustered))
```

### Bread-matrix implementation

In statistics:
$$
\text{Var}(\hat{\beta}) \approx [-H(\hat{\beta})]^{-1}
$$

In this package, bread is defined as:
$$
B = n \cdot \text{Var}(\hat{\beta}) = n \cdot [-H(\hat{\beta})]^{-1}
$$

So:
$$
[-H(\hat{\beta})]^{-1} = \frac{B}{n}
$$

Substituting into the one-step update:
$$
\beta^* \approx \hat{\beta} + \frac{B}{n} \cdot U_{\text{star}}
= \hat{\beta} + \frac{1}{n}(B \cdot U_{\text{star}})
$$

We compute $(B \cdot U_{\text{star}})$ first (matrix times vector), then divide
by $n$.

``` r
beta_draws[i, ] <- beta_hat + as.vector((bread %*% U_star) / n_obs)
```

------------------------------------------------------------------------

## 4. Usage in G-Computation

Once we have the perturbed parameters $\beta^*$, we use them to generate the
population-average (marginal) SOPs.

### Step A: Predict

For each patient $i$, we calculate their individual-level SOPs given the new
parameters:
$$
\text{SOP}_i(t, s) = f(\text{Data}_i, \beta^*)
$$

``` r
# Fast Path (Pre-calculated components)
sops_array <- markov_msm_run(components, Gamma_i, times, absorb)

# Slow Path (Full prediction)
sops_array <- soprob_markov(object = model_i, data = newdata_pred, ...)
```

### Step B: Weighted Average

We then average these individual probabilities across the population.
Crucially, we use the same weights $w_g$ that we used to perturb the
coefficients to account for sampling uncertainty:
$$
\overline{\text{SOP}}(t, s) =
\frac{\sum_{i=1}^N w_i \cdot \text{SOP}_i(t, s)}{\sum_{i=1}^N w_i}
$$

``` r
# Normalizing to a simplex
weights <- weights / weight_sum

# Weighted marginalization (Dot product of probabilities and weights)
avg_sops_mat <- matrix(as.vector(sops_cf_mat %*% weights), ...)
```

**Why use the same weights?** The bootstrap samples patients with the weights
$w_g$. To maintain consistency, we must:

1. Perturb coefficients using weights $w_g$ (reflecting sampling uncertainty)
2. Average patient predictions using the *same* weights $w_g$ (reflecting the
   "sampled" population)

By normalizing by $\sum_{i=1}^N w_i$ (the simplex), we ensure the final
marginalized estimate remains a valid probability between 0 and 1.

------------------------------------------------------------------------

## Appendix: First-Order Taylor Justification

Instead of fully refitting the weighted likelihood for each draw, we linearize
$U^*(\beta)$ around $\hat{\beta}$:
$$
U^*(\beta^*) \approx U^*(\hat{\beta}) +
\frac{\partial U^*(\hat{\beta})}{\partial \beta} (\beta^* - \hat{\beta}) =
\mathbf{0}
$$

Let $\Delta\beta = (\beta^* - \hat{\beta})$. Rearranging gives:
$$
\Delta\beta \approx -
\left[ \frac{\partial U^*(\hat{\beta})}{\partial \beta} \right]^{-1}
U^*(\hat{\beta})
$$

The derivative term is the weighted Hessian $\sum_g w_g H_g$. Since
$E[w_g]=1$, we approximate it with the original Hessian and recover the
one-step update used in the main text.
