---
title: "Score Bootstrap Intuition in a One-Parameter Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Score Bootstrap Intuition in a One-Parameter Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 300
)
```

## Goal

This vignette builds score-bootstrap intuition in progressively harder steps.
We start with the simplest case (a single intercept parameter), then move to
repeated-simulation checks.

Roadmap:

1. On one fixed dataset: fit `VGAM::vglm(y ~ 1)`, inspect log-likelihood
   geometry, perturb score contributions, and apply the one-step update.
2. On that same dataset: compare 1000 score-bootstrap draws to 1000
   nonparametric bootstrap draws.
3. Across many simulated datasets under the null: assess calibration for
   intercept inference, then repeat for the `tx` coefficient in a simple
   Markov transition model.

## Setup

```{r setup, message=FALSE}
library(markov.misc)
library(VGAM)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)
library(furrr)
library(future.callr)

SEED <- 20260216
N <- 100
P_TRUE <- 0.6
N_DRAWS <- 1000

set.seed(SEED)
dat <- tibble(
  id = seq_len(N),
  y_bin = rbinom(N, size = 1, prob = P_TRUE),
  y = ordered(y_bin, levels = c(0, 1))
)

fit <- vglm(
  y ~ 1,
  family = cumulative(reverse = TRUE),
  data = dat
)

# One-observation-per-cluster to mirror the score-bootstrap machinery
# while intentionally ignoring clustering structure.
fit_rb <- robcov_vglm(fit, cluster = dat$id)

beta_hat <- as.numeric(stats::coef(fit)[1])
scores <- as.numeric(fit_rb$scores[, 1])
bread <- as.numeric(fit_rb$bread[1, 1])
n_obs <- fit_rb$n
v_hat <- bread / n_obs

stopifnot(length(stats::coef(fit)) == 1)
stopifnot(abs(sum(scores)) < 1e-8)
stopifnot(isTRUE(all.equal(v_hat, as.numeric(stats::vcov(fit)[1, 1]), tolerance = 1e-10)))

y_num <- dat$y_bin
```

For this fixed dataset, the fitted intercept corresponds to:

```{r}
tibble(
  quantity = c("beta_hat", "plogis(beta_hat)", "sample mean(y)"),
  value = c(beta_hat, plogis(beta_hat), mean(y_num))
)
```

## Log-Likelihood Geometry in 1D

For Bernoulli outcomes with one intercept:

- log-likelihood: $\ell(\beta) = \sum_i \{y_i \beta - \log(1 + e^\beta)\}$,
- score: $U(\beta) = \sum_i (y_i - \text{logit}^{-1}(\beta))$,
- Hessian: $H(\beta) = -n\,p(\beta)\{1-p(\beta)\}$.

```{r ll-score-hess}
ll_fn <- function(beta, y) {
  sum(y * beta - log1p(exp(beta)))
}

score_fn <- function(beta, y) {
  sum(y - plogis(beta))
}

hess_fn <- function(beta, n) {
  p <- plogis(beta)
  -n * p * (1 - p)
}

beta_grid <- seq(beta_hat - 4, beta_hat + 4, length.out = 500)
geom_df <- tibble(
  beta = beta_grid,
  loglik = vapply(beta_grid, ll_fn, numeric(1), y = y_num),
  score = vapply(beta_grid, score_fn, numeric(1), y = y_num),
  hess = vapply(beta_grid, hess_fn, numeric(1), n = n_obs)
)

p_ll <- ggplot(geom_df, aes(beta, loglik)) +
  geom_line(linewidth = 0.8, color = "#2C3E50") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "Log-likelihood",
    x = expression(beta),
    y = expression(ell(beta))
  ) +
  theme_minimal()

p_score <- ggplot(geom_df, aes(beta, score)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.8, color = "#2874A6") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "First derivative (Score)",
    x = expression(beta),
    y = expression(U(beta))
  ) +
  theme_minimal()

p_hess <- ggplot(geom_df, aes(beta, hess)) +
  geom_line(linewidth = 0.8, color = "#117864") +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "Second derivative (Hessian)",
    x = expression(beta),
    y = expression(H(beta))
  ) +
  theme_minimal()

(p_ll / p_score / p_hess) +
  plot_annotation(title = "Likelihood Geometry for a Single Intercept")
```

At the MLE (`beta_hat`), the score is approximately zero.

## Score Contributions and Exponential Reweighting

Keeping this dataset fixed, we now perturb the score equation the same way the
score bootstrap does.

`robcov_vglm()` gives observation-level score contributions $s_i$ at
$\hat\beta$. In this one-parameter case, these are scalar values.

For one score-bootstrap draw:

- sample weights $w_i \sim \text{Exp}(1)$,
- center them as $u_i = w_i - 1$,
- compute perturbed score at the MLE:
  $$
  U^*(\hat\beta) = \sum_i u_i s_i.
  $$

```{r single-perturbation}
set.seed(SEED + 1)
w_i <- rexp(n_obs, rate = 1)
u_i <- w_i - 1

contrib_df <- tibble(
  obs = seq_len(n_obs),
  score_i = scores,
  centered_weight = u_i,
  weighted_score_i = u_i * score_i
)

score_totals <- tibble(
  type = c("Original total score", "Perturbed total score"),
  value = c(sum(contrib_df$score_i), sum(contrib_df$weighted_score_i))
)

knitr::kable(score_totals, digits = 4)
```

```{r contribution-viz}
p_contrib <- ggplot(contrib_df, aes(obs, weighted_score_i)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_linerange(aes(ymin = 0, ymax = weighted_score_i), color = "#2874A6", alpha = 0.6) +
  labs(
    title = "Perturbed contribution per observation",
    x = "Observation",
    y = expression((w[i] - 1) * s[i])
  ) +
  theme_minimal()

p_contrib
```

## How the Total Score Curve Changes

The key geometric effect is that exponential reweighting tilts the score curve
and moves its root.

The unweighted score curve is:
$$
U(\beta) = \sum_i (y_i - \text{logit}^{-1}(\beta)).
$$

With one exponential draw, the weighted score curve is:
$$
U_w(\beta) = \sum_i w_i (y_i - \text{logit}^{-1}(\beta)).
$$

At $\hat\beta$, $U(\hat\beta)=0$, while generally $U_w(\hat\beta)\neq 0$.

```{r score-shift-viz}
score_curve_df <- tibble(
  beta = beta_grid,
  unweighted = vapply(beta_grid, score_fn, numeric(1), y = y_num),
  weighted = vapply(
    beta_grid,
    function(b) sum(w_i * (y_num - plogis(b))),
    numeric(1)
  )
) |>
  pivot_longer(cols = c(unweighted, weighted), names_to = "curve", values_to = "score")

overview <- ggplot(score_curve_df, aes(beta, score, color = curve)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "grey35") +
  scale_color_manual(
    values = c(unweighted = "#2C3E50", weighted = "#C0392B"),
    labels = c(unweighted = "U(beta)", weighted = "U_w(beta)")
  ) +
  # coord_cartesian(xlim = c(0.5, 1.25), ylim = c(-5, 5)) +
  labs(
    title = "Score curve before and after reweighting",
    x = expression(beta),
    y = "Score",
    color = NULL
  ) +
  theme_minimal()

zoomed <- ggplot(score_curve_df, aes(beta, score, color = curve)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = beta_hat, linetype = "dashed", color = "grey35") +
  scale_color_manual(
    values = c(unweighted = "#2C3E50", weighted = "#C0392B"),
    labels = c(unweighted = "U(beta)", weighted = "U_w(beta)")
  ) +
  coord_cartesian(xlim = c(0.5, 1.25), ylim = c(-5, 5)) +
  labs(
    title = "Score curve before and after reweighting (zoomed)",
    x = expression(beta),
    y = "Score",
    color = NULL
  ) +
  theme_minimal()

overview / zoomed
```

## One-Step Update as a Tangent Approximation

Given that shifted score curve, the package uses a one-step Newton-style update
to approximate the new root.

In 1D, the one-step score-bootstrap update used in the package is:
$$
\beta^* \approx \hat\beta + V(\hat\beta) \, U^*(\hat\beta),
$$


where $V(\hat\beta)=(-H(\hat\beta))^{-1}$, and in package terms
$V(\hat\beta) = \texttt{bread}/n$.

```{r tangent-viz}
U_star <- sum(contrib_df$weighted_score_i)
H_hat <- hess_fn(beta_hat, n_obs)
beta_one_step <- beta_hat + v_hat * U_star

weighted_score_fn <- function(beta) {
  sum(w_i * (y_num - plogis(beta)))
}

find_root <- function(f, center, step = 2, max_iter = 25) {
  left <- center - step
  right <- center + step
  f_left <- f(left)
  f_right <- f(right)
  iter <- 0

  while (sign(f_left) == sign(f_right) && iter < max_iter) {
    step <- step * 1.5
    left <- center - step
    right <- center + step
    f_left <- f(left)
    f_right <- f(right)
    iter <- iter + 1
  }

  if (sign(f_left) == sign(f_right)) {
    return(NA_real_)
  }

  stats::uniroot(f, c(left, right))$root
}

beta_exact <- find_root(weighted_score_fn, center = beta_hat)
stopifnot(is.finite(beta_exact))

tangent_df <- tibble(
  beta = beta_grid,
  weighted_score = vapply(beta_grid, weighted_score_fn, numeric(1)),
  tangent = U_star + H_hat * (beta - beta_hat)
)

ggplot(tangent_df, aes(beta)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_line(aes(y = weighted_score, color = "Weighted score curve"), linewidth = 1) +
  geom_line(aes(y = tangent, color = "Tangent at beta_hat"), linewidth = 1, linetype = "dotted") +
  geom_vline(aes(xintercept = beta_hat, color = "beta_hat"), linetype = "dotdash") +
  geom_vline(aes(xintercept = beta_one_step, color = "One-step beta*"), linetype = "dotdash") +
  geom_vline(aes(xintercept = beta_exact, color = "Exact weighted-MLE root"), linetype = "dotdash") +
  scale_color_manual(
    values = c(
      "Weighted score curve" = "#2C3E50",
      "Tangent at beta_hat" = "#7D3C98",
      "beta_hat" = "#1E8449",
      "One-step beta*" = "#D35400",
      "Exact weighted-MLE root" = "#C0392B"
    )
  ) +
  coord_cartesian(xlim = c(0.7, 1.0), ylim = c(-5, 5)) +
  labs(
    title = "One-step Newton update from the tangent line",
    x = expression(beta),
    y = "Weighted score",
    color = NULL
  ) +
  theme_minimal()
```

```{r one-step-comparison}
tibble(
  quantity = c("beta_hat", "U_star(beta_hat)", "beta_one_step", "beta_exact"),
  value = c(beta_hat, U_star, beta_one_step, beta_exact)
) |>
  knitr::kable(digits = 5)
```

## 1000 Score-Bootstrap Draws vs 1000 Nonparametric Bootstrap Draws

We now move from a single perturbation to many draws and compare distributions
on the same observed dataset.

```{r draw-generators}
draw_score_boot <- function(n_draws, beta_hat, v_hat, scores) {
  n_obs <- length(scores)
  u_mat <- matrix(rexp(n_draws * n_obs, rate = 1) - 1, nrow = n_draws, ncol = n_obs)
  U_star <- as.vector(u_mat %*% scores)
  beta_hat + v_hat * U_star
}

draw_nonparam_boot <- function(n_draws, dat) {
  out <- numeric(n_draws)
  n <- nrow(dat)

  for (i in seq_len(n_draws)) {
    idx <- sample.int(n, size = n, replace = TRUE)
    dat_b <- dat[idx, , drop = FALSE]
    fit_b <- vglm(y ~ 1, family = cumulative(reverse = TRUE), data = dat_b)
    out[i] <- as.numeric(stats::coef(fit_b)[1])
  }

  out
}
```

```{r run-draws}
set.seed(SEED + 2)
t_score <- system.time({
  beta_score <- draw_score_boot(
    n_draws = N_DRAWS,
    beta_hat = beta_hat,
    v_hat = v_hat,
    scores = scores
  )
})["elapsed"]

set.seed(SEED + 3)
t_nonparam <- system.time({
  beta_nonparam <- draw_nonparam_boot(
    n_draws = N_DRAWS,
    dat = dat
  )
})["elapsed"]

stopifnot(length(beta_score) == N_DRAWS, length(beta_nonparam) == N_DRAWS)
stopifnot(all(is.finite(beta_score)), all(is.finite(beta_nonparam)))
```

```{r draw-summary}
summarize_draws <- function(x, method) {
  tibble(
    method = method,
    mean = mean(x),
    sd = stats::sd(x),
    q025 = as.numeric(stats::quantile(x, 0.025)),
    q50 = as.numeric(stats::quantile(x, 0.5)),
    q95 = as.numeric(stats::quantile(x, 0.95)),
    q975 = as.numeric(stats::quantile(x, 0.975))
  )
}

summary_tbl <- bind_rows(
  summarize_draws(beta_score, "Score bootstrap"),
  summarize_draws(beta_nonparam, "Nonparametric bootstrap")
)

runtime_tbl <- tibble(
  method = c("Score bootstrap", "Nonparametric bootstrap"),
  elapsed_seconds = c(as.numeric(t_score), as.numeric(t_nonparam))
)

knitr::kable(summary_tbl, digits = 4, caption = "Distribution summary of intercept draws")
knitr::kable(runtime_tbl, digits = 3, caption = "Runtime for 1000 draws")
```

```{r distribution-comparison}
draws_df <- bind_rows(
  tibble(beta = beta_score, method = "Score bootstrap"),
  tibble(beta = beta_nonparam, method = "Nonparametric bootstrap")
)

p_density <- ggplot(draws_df, aes(beta, fill = method, color = method)) +
  geom_density(alpha = 0.20, linewidth = 0.9) +
  labs(
    title = "Intercept draw distributions",
    x = expression(beta),
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme_minimal()

probs <- seq(0.01, 0.99, by = 0.01)
qq_df <- tibble(
  nonparametric = as.numeric(stats::quantile(beta_nonparam, probs = probs)),
  score_bootstrap = as.numeric(stats::quantile(beta_score, probs = probs))
)

p_qq <- ggplot(qq_df, aes(nonparametric, score_bootstrap)) +
  geom_point(alpha = 0.7, size = 1.4, color = "#2C3E50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "QQ comparison",
    x = "Nonparametric bootstrap quantiles",
    y = "Score bootstrap quantiles"
  ) +
  theme_minimal()

p_density | p_qq
```

## Repeated-Simulation Check: Intercept Inference

The previous section compared draw distributions for one dataset. Now we assess
whether score-bootstrap intervals are calibrated across many independently
simulated datasets.

We use a null data-generating process with $p = 0.5$, so the true intercept
is $\beta_{\text{true}} = 0$. Type I error is the proportion of simulations
where a 95% interval excludes 0.

```{r simulation-settings}
N_SIM <- 1000
N_SIM_OBS <- 100
P_SIM <- 0.5
BETA_TRUE <- qlogis(P_SIM)
B_SIM_DRAWS <- 1000
ALPHA <- 0.05
SIM_WORKERS <- 6
```

```{r simulation-helper}
run_one_sim <- function(sim_id, n_obs, p_true, n_draws) {
  dat_sim <- tibble(
    id = seq_len(n_obs),
    y_bin = rbinom(n_obs, size = 1, prob = p_true),
    y = ordered(y_bin, levels = c(0, 1))
  )

  fit_sim <- vglm(
    y ~ 1,
    family = cumulative(reverse = TRUE),
    data = dat_sim
  )
  fit_sim_rb <- robcov_vglm(fit_sim, cluster = dat_sim$id)

  beta_hat_sim <- as.numeric(stats::coef(fit_sim)[1])
  scores_sim <- as.numeric(fit_sim_rb$scores[, 1])
  v_hat_sim <- as.numeric(fit_sim_rb$bread[1, 1] / fit_sim_rb$n)

  beta_score_sim <- draw_score_boot(
    n_draws = n_draws,
    beta_hat = beta_hat_sim,
    v_hat = v_hat_sim,
    scores = scores_sim
  )

  ci_low <- as.numeric(stats::quantile(beta_score_sim, 0.025))
  ci_high <- as.numeric(stats::quantile(beta_score_sim, 0.975))
  draw_mean <- mean(beta_score_sim)

  tibble(
    sim_id = sim_id,
    method = "Score bootstrap",
    beta_hat = beta_hat_sim,
    mle_bias = beta_hat_sim - BETA_TRUE,
    draw_mean = draw_mean,
    bias = draw_mean - BETA_TRUE,
    ci_low = ci_low,
    ci_high = ci_high,
    ci_width = ci_high - ci_low,
    covered = (ci_low <= BETA_TRUE) && (ci_high >= BETA_TRUE),
    reject_h0 = (ci_low > 0) || (ci_high < 0)
  )
}
```

```{r run-simulations}
run_sim_map <- function() {
  furrr::future_map_dfr(
    seq_len(N_SIM),
    ~run_one_sim(.x, n_obs = N_SIM_OBS, p_true = P_SIM, n_draws = B_SIM_DRAWS),
    .options = furrr::furrr_options(seed = TRUE),
    .progress = TRUE
  )
}

future::plan(future.callr::callr, workers = SIM_WORKERS)

set.seed(SEED + 4)
t_sim <- system.time({
  sim_results <- run_sim_map()
})["elapsed"]

future::plan(future::sequential)
```

```{r sim-summary}
sim_summary <- sim_results |>
  summarise(
    method = "Score bootstrap",
    mean_mle_bias = mean(mle_bias),
    mean_draw_bias = mean(bias),
    empirical_sd_draw_mean = stats::sd(draw_mean),
    coverage = mean(covered),
    type1_error = mean(reject_h0),
    avg_ci_width = mean(ci_width)
  )

sim_summary <- sim_summary |>
  mutate(
    mcse_coverage = sqrt((coverage * (1 - coverage)) / N_SIM),
    mcse_type1 = sqrt((type1_error * (1 - type1_error)) / N_SIM)
  )

# sim_runtime <- tibble(
#   setting = paste0(
#     N_SIM,
#     " simulations (",
#     B_SIM_DRAWS,
#     " score-bootstrap draws per simulation)"
#   ),
#   elapsed_seconds = as.numeric(t_sim)
# )

knitr::kable(sim_summary, digits = 4, caption = "Simulation performance summary")
```

```{r sim-plot}
nominal_df <- tibble(
  metric = c("Coverage", "Type I error"),
  estimate = c(sim_summary$coverage, sim_summary$type1_error),
  mcse = c(sim_summary$mcse_coverage, sim_summary$mcse_type1),
  nominal = c(1 - ALPHA, ALPHA)
) |>
  mutate(
    low = pmax(0, estimate - 1.96 * mcse),
    high = pmin(1, estimate + 1.96 * mcse)
  )

p_nominal <- ggplot(nominal_df, aes(metric, estimate)) +
  geom_col(width = 0.55, fill = "#2C3E50", alpha = 0.9) +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.12) +
  geom_point(aes(y = nominal), color = "#C0392B", size = 2.8) +
  geom_segment(
    aes(x = as.numeric(factor(metric)) - 0.22, xend = as.numeric(factor(metric)) + 0.22, y = nominal, yend = nominal),
    linetype = "dashed",
    color = "#C0392B"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Nominality Check (Estimate with MCSE)",
    x = NULL,
    y = "Probability"
  ) +
  theme_minimal()

p_bias <- ggplot(sim_results, aes(x = bias)) +
  geom_histogram(bins = 18, fill = "#2874A6", color = "white", alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#C0392B") +
  labs(
    title = "Simulation-level Bias (Draw Mean - True beta)",
    x = "Bias",
    y = "Count"
  ) +
  theme_minimal()

p_nominal | p_bias
```

Because this is a null scenario ($\beta_{\text{true}}=0$), type I error
should be close to $1-$coverage, and average bias should be near zero.

## Repeated-Simulation Check: `tx` Null Coverage in a Markov Model

Finally, we keep the same inferential target (null calibration) but move from
the intercept-only model to a simple proportional-odds transition model. The
focus is nominal coverage for the treatment coefficient `tx` when the true
effect is zero.

This is a **coefficient-level** target ($\beta_{\text{tx}} = 0$); it is not a
population-average treatment effect.

### Simulation Settings

```{r tx-coverage-settings}
N_PATIENTS_TX <- 100
FU_TX <- 30
N_SIM_TX <- 500
N_DRAWS_TX <- 1000
ALPHA_TX <- 0.05
SIM_WORKERS_TX <- 6
BETA_TX_TRUE <- 0
```

### One-Step Score-Bootstrap Draws for Coefficients

```{r tx-coverage-helper}
draw_score_boot_coefs <- function(model, n_draws) {
  required <- c("coefficients", "bread", "scores", "cluster", "n")
  missing_required <- required[vapply(
    required,
    function(x) is.null(model[[x]]),
    logical(1)
  )]
  if (length(missing_required) > 0) {
    stop(
      "robcov_vglm model is missing required components: ",
      paste(missing_required, collapse = ", ")
    )
  }

  beta_hat <- model$coefficients
  bread <- model$bread
  scores <- model$scores
  cluster <- model$cluster
  n_obs <- model$n

  if (!is.matrix(scores)) {
    stop("`scores` must be a matrix.")
  }

  cluster_chr <- as.character(cluster)
  cluster_ids <- unique(cluster_chr)
  cluster_fac <- factor(cluster_chr, levels = cluster_ids)
  scores_clustered <- rowsum(scores, cluster_fac, reorder = FALSE)

  n_clusters <- nrow(scores_clustered)
  p <- length(beta_hat)

  beta_draws <- matrix(NA_real_, nrow = n_draws, ncol = p)
  colnames(beta_draws) <- names(beta_hat)

  for (i in seq_len(n_draws)) {
    u_cluster <- rexp(n_clusters, rate = 1) - 1
    U_star <- as.vector(crossprod(u_cluster, scores_clustered))
    beta_draws[i, ] <- beta_hat + as.vector((bread %*% U_star) / n_obs)
  }

  beta_draws
}

run_one_tx_sim <- function(sim_id, n_patients, fu, n_draws, alpha) {
  tryCatch(
    {
      raw_data <- sim_trajectories_brownian(
        n_patients = n_patients,
        follow_up_time = fu,
        treatment_prob = 0.5,
        absorbing_state = 6,
        mu_treatment_effect = 0
      )

      age_lookup <- tibble(
        id = sort(unique(raw_data$id)),
        age = rnorm(dplyr::n_distinct(raw_data$id), mean = 65, sd = 12)
      )
      raw_data <- raw_data |>
        left_join(age_lookup, by = "id")

      data <- prepare_markov_data(raw_data)
      time_spl <- rms::rcs(data$time, 4)
      data <- data |>
        mutate(
          time_lin = as.vector(time_spl[, 1]),
          time_nlin_1 = as.vector(time_spl[, 2]),
          time_nlin_2 = as.vector(time_spl[, 3])
        )

      m_simple <- vglm(
        ordered(y) ~ (time_lin + time_nlin_1 + time_nlin_2) + tx + age + yprev + time_lin:yprev,
        family = cumulative(reverse = TRUE, parallel = TRUE),
        data = data
      )
      m_simple_rb <- robcov_vglm(m_simple, cluster = data$id)

      beta_draws <- draw_score_boot_coefs(m_simple_rb, n_draws = n_draws)
      if (!"tx" %in% colnames(beta_draws)) {
        stop("Coefficient `tx` was not found in the fitted model.")
      }

      tx_hat <- as.numeric(m_simple_rb$coefficients["tx"])
      tx_draws <- as.numeric(beta_draws[, "tx"])
      if (!all(is.finite(tx_draws))) {
        stop("Non-finite tx score-bootstrap draws.")
      }

      ci_low <- as.numeric(stats::quantile(tx_draws, alpha / 2))
      ci_high <- as.numeric(stats::quantile(tx_draws, 1 - alpha / 2))
      draw_mean <- mean(tx_draws)

      tibble(
        sim_id = sim_id,
        success = TRUE,
        error_msg = NA_character_,
        tx_hat = tx_hat,
        draw_mean = draw_mean,
        draw_bias = draw_mean - BETA_TX_TRUE,
        ci_low = ci_low,
        ci_high = ci_high,
        ci_width = ci_high - ci_low,
        covered = (ci_low <= BETA_TX_TRUE) && (ci_high >= BETA_TX_TRUE),
        reject_h0 = (ci_low > BETA_TX_TRUE) || (ci_high < BETA_TX_TRUE)
      )
    },
    error = function(e) {
      tibble(
        sim_id = sim_id,
        success = FALSE,
        error_msg = conditionMessage(e),
        tx_hat = NA_real_,
        draw_mean = NA_real_,
        draw_bias = NA_real_,
        ci_low = NA_real_,
        ci_high = NA_real_,
        ci_width = NA_real_,
        covered = NA,
        reject_h0 = NA
      )
    }
  )
}
```

### Run Coverage Simulation

```{r tx-coverage-run}
run_tx_sim_map <- function() {
  furrr::future_map_dfr(
    seq_len(N_SIM_TX),
    ~run_one_tx_sim(
      sim_id = .x,
      n_patients = N_PATIENTS_TX,
      fu = FU_TX,
      n_draws = N_DRAWS_TX,
      alpha = ALPHA_TX
    ),
    .options = furrr::furrr_options(seed = TRUE),
    .progress = TRUE
  )
}

future::plan(future.callr::callr, workers = SIM_WORKERS_TX)

set.seed(SEED + 876)
t_tx <- system.time({
  tx_results <- run_tx_sim_map()
})["elapsed"]

future::plan(future::sequential)
```

### Summary

```{r tx-coverage-summary}
tx_results_ok <- tx_results |>
  filter(success)

stopifnot(nrow(tx_results_ok) > 0)

tx_summary <- tx_results_ok |>
  summarise(
    n_attempted = N_SIM_TX,
    n_successful = n(),
    failure_rate = 1 - (n() / N_SIM_TX),
    mean_tx_hat = mean(tx_hat),
    mean_draw_bias = mean(draw_bias),
    coverage = mean(covered),
    type1_error = mean(reject_h0),
    avg_ci_width = mean(ci_width)
  ) |>
  mutate(
    mcse_coverage = sqrt((coverage * (1 - coverage)) / n_successful),
    mcse_type1 = sqrt((type1_error * (1 - type1_error)) / n_successful)
  )

tx_runtime <- tibble(
  setting = paste0(
    N_SIM_TX,
    " simulations (",
    N_DRAWS_TX,
    " score-bootstrap draws per simulation)"
  ),
  n_attempted = N_SIM_TX,
  n_successful = tx_summary$n_successful,
  elapsed_seconds = as.numeric(t_tx)
)

knitr::kable(tx_summary, digits = 4, caption = "Null-coverage summary for the tx coefficient")
knitr::kable(tx_runtime, digits = 3, caption = "Runtime for tx coverage simulation")
```

```{r tx-coverage-errors}
# tx_failures <- tx_results |>
#   filter(!success) |>
#   count(error_msg, sort = TRUE) |>
#   slice_head(n = 5)

# if (nrow(tx_failures) > 0) {
#   knitr::kable(tx_failures, caption = "Top simulation failure messages")
# }
```

```{r tx-coverage-plot}
tx_nominal_df <- tibble(
  metric = c("Coverage", "Type I error"),
  estimate = c(tx_summary$coverage, tx_summary$type1_error),
  mcse = c(tx_summary$mcse_coverage, tx_summary$mcse_type1),
  nominal = c(1 - ALPHA_TX, ALPHA_TX)
) |>
  mutate(
    low = pmax(0, estimate - 1.96 * mcse),
    high = pmin(1, estimate + 1.96 * mcse)
  )

ggplot(tx_nominal_df, aes(metric, estimate)) +
  geom_col(width = 0.55, fill = "#2C3E50", alpha = 0.9) +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.12) +
  geom_point(aes(y = nominal), color = "#C0392B", size = 2.8) +
  geom_segment(
    aes(
      x = as.numeric(factor(metric)) - 0.22,
      xend = as.numeric(factor(metric)) + 0.22,
      y = nominal,
      yend = nominal
    ),
    linetype = "dashed",
    color = "#C0392B"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Nominality Check for tx Coefficient (Estimate with MCSE)",
    x = NULL,
    y = "Probability"
  ) +
  theme_minimal()
```
