---
title: "Second-Order SOPs with orm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Second-Order SOPs with orm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

For a second-order transition model, the transition probability depends on the
two previous states. The recursion therefore tracks a joint distribution over
the last two states, then marginalizes it back to SOPs.

This example uses `orm_markov()` with `yprev` and `ypprev`.

## Data

```{r setup}
library(markov.misc)
library(ggplot2)
library(rms)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260529
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_brownian(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  mu_treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
model_data <- model_data[order(model_data$id, model_data$time), ]
model_data$ypprev <- ave(
  model_data$yprev,
  model_data$id,
  FUN = function(x) c(x[1], head(x, -1))
)
model_data$ypprev <- factor(model_data$ypprev, levels = levels(model_data$yprev))

head(model_data[c("id", "time", "tx", "ypprev", "yprev", "y")])
```

## Second-Order Recursion

For a second-order transition model, the SOP for state $l$ at time $t$ is

$$
S_{it}(l) =
  \Pr(Y_{it} = l \mid \mathbf{X}_{i1}, \ldots, \mathbf{X}_{it},
  Y_{i,-1}, Y_{i0}),
\qquad t \geq 1,
$$

where $Y_{i,-1}$ and $Y_{i0}$ are the two observed initial states. The
conditional transition probability at time $t$ is

$$
T_{it}(l \mid h, k) =
  \Pr(Y_{it} = l \mid Y_{i,t-2} = h, Y_{i,t-1} = k, \mathbf{X}_{it}).
$$

These probabilities are predicted directly from the fitted ordinal transition
model by setting `ypprev` to $h$, `yprev` to $k$, and `time` to $t$.
Because the next state depends on two previous states, the recursion tracks the
joint probability matrix of the last two outcomes:

$$
J_{it}(k,l) =
  \Pr(Y_{i,t-1} = k, Y_{it} = l \mid \mathbf{X}_i, Y_{i,-1}, Y_{i0}).
$$

Equivalently, one can initialize
$J_{i0}(h,k) = \mathbb{1}\{h = Y_{i,-1}, k = Y_{i0}\}$ and update from there.
The code
uses the same recursion, with a small computational shortcut: it first predicts
the distribution of $Y_{i1}$ and stores

$$
J_{i1}(k,l) =
  \mathbb{1}\{k = Y_{i0}\} T_{i1}(l \mid Y_{i,-1}, Y_{i0}).
$$

For later times, the joint distribution is updated by marginalizing over the
older state $h$:

$$
J_{it}(k,l) =
  \sum_{h=1}^{K}
    J_{i,t-1}(h,k) T_{it}(l \mid h,k),
\qquad t = 2,\ldots,d.
$$

The SOP is then the column margin of the joint matrix:

$$
S_{it}(l) =
  \sum_{k=1}^{K} J_{it}(k,l).
$$

In matrix/tensor notation, define the transition tensor
$\mathcal{T}_{it}$ with entries
$\mathcal{T}_{it,hkl} = T_{it}(l \mid h,k)$. The update is a tensor
contraction over the oldest-state index:

$$
J_{it,kl} =
  \sum_{h=1}^{K}
    J_{i,t-1,hk} \mathcal{T}_{it,hkl},
\qquad
s_{it,l} =
  \sum_{k=1}^{K} J_{it,kl}.
$$

Batched over patients, this is the elementwise tensor calculation used by
`soprob_markov()`:

$$
\mathbf{J}_{t,kl} =
  \sum_{h=1}^{K}
    \mathbf{J}_{t-1,hk} \odot \mathbf{T}_{t,hkl},
\qquad
\mathbf{S}_{t,l} =
  \sum_{k=1}^{K} \mathbf{J}_{t,kl}.
$$

If $k$ is absorbing, the implementation skips model prediction from $k$ and
carries the joint mass forward to $J_{it}(k,k)$. Thus for absorbing
$k \in \mathcal{A}$,

$$
J_{it}(k,k) \leftarrow
  J_{it}(k,k) + \sum_{h=1}^{K} J_{i,t-1}(h,k).
$$

`avg_sops(..., p2varname = "ypprev")` uses this array recursion instead of
simulating individual trajectories.

## Fit And Predict

```{r fit-model}
dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_second <- orm_markov(
  y ~ tx + rms::rcs(time, 4) + yprev + ypprev + time%ia%yprev + time%ia%ypprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

fit_second
```

```{r avg-sops}
sop_second <- avg_sops(
  fit_second,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  p2varname = "ypprev"
) |> 
  inferences(n_sim = 50)

head(sop_second)
```

```{r plot-sops}
plot_sops(
  sop_second,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Model Diagnostics

These diagnostics compare empirical trajectory features with the corresponding
features implied by the second-order transition model. The model-based plots
pass `p2varname = "ypprev"` so the deterministic recursions condition on both
previous states.

```{r diagnostics-transitions-observed}
diagnostic_times <- c(1, 3, 5, 7, 11, 13)

plot_transitions(
  raw_data,
  times = diagnostic_times,
  ylevels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model}
plot_transitions(
  fit_second,
  times = diagnostic_times,
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  p2varname = "ypprev",
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed}
plot_correlation(
  raw_data,
  times = seq_len(FOLLOW_UP),
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model}
plot_correlation(
  fit_second,
  times = seq_len(FOLLOW_UP),
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  p2varname = "ypprev",
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed}
plot_variogram(
  raw_data,
  times = seq_len(FOLLOW_UP)
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model}
plot_variogram(
  fit_second,
  times = seq_len(FOLLOW_UP),
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  p2varname = "ypprev"
) +
  labs(title = "Model-based correlation variogram")
```

## Contrast

```{r}
c <- avg_comparisons(
  fit_second,
  metric = "time_in_state",
  states = list(1:2, 3:5, 6:7, 8),
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  ylevels = fit_second$yunique,
  absorb = ABSORB,
  p2varname = "ypprev"
) |> 
  inferences(n_sim = 100)

plot_comparisons(c)
```
